---
title: "01 Phyloseq Adult"
author: "Madeline Krieger"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: "cosmo"
    code_folding: "show"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      error = FALSE,
                      fig.path = "../figures/01_adult/",
                      fig.keep = "all",
                      dev = c("png"),
                      cache = FALSE,
                      # Do not use lazy loading which fails for large objects
                      cache.lazy = FALSE)
```


# 1.0 Load and Define

## 1.1 Load packages and presets
### 1.1.1 Packages
```{r 1.1_load-packages}
library("pacman")
pacman::p_load("rmarkdown", "tidyverse", "phyloseq", "knitr","kableExtra", "here", "plyr", "ggpubr", "microViz", "microbiome", "pheatmap", "vegan", "reshape2", "magrittr", "microshades", "pheatmap","vegan", "data.table", "Polychrome", "fantaxtic","cetcolor", "cowplot", "MicrobiomeStat", "randomForest", "caret", "mlbench", "MLmetrics", "mia", "here", "patchwork", "digest", "ANCOMBC", "Maaslin2", "microbiomeMarker", "fantaxtic", "topicmodels", "ldatuning", "umap", "tidytext", "cluster")

here::i_am("01_adult_cohort.Rmd")
``` 


### 1.1.2 My packages
```{r}
source("MK_Microbiome_Functions.R")
```



### 1.1.2 Colors and presents
```{r}
colors_all= c("Abscess" = "#FF495C", "Plaque"="#083D77")


core_colors = c("#FAAA00", "#3399FF","#F76F8E","#083D77","#B8D4E3", "#FF495C","#477071", '#03CEA4',  "#5F00BA", "#BDAC9E", "white", "#FFD900")

maaslin2_colors= c("CLR_LOG" = "#2c45b5", "CLR_NONE"="#86B8FD", 
                   "CSS_LOG" = "#FF495C", "CSS_NONE"="#F2929A",
                   "TMM_LOG" = "#F4B701", "TMM_NONE"="#FEE59A",
                   "TSS_LOG" = "#8FD694", "TSS_NONE"="#D2EED4",
                   "TSS_LOGIT" = "#354F52", "TSS_AST"="#CEDDDF")

pal20 <- createPalette(22, c("#F76F8E","#03CEA4", "#083D77"), range = c(30, 80))
pal20 <- unname(as.list(pal20))

resolution <- c("Phylum", "Class", "Order", "Family", "Genus", "Species")

set.seed(1234)
date = Sys.Date()
```


## 1.2 Load data and process

### 1.2.1 Load Data from 00
```{r}
# Load the saved Rds file
loaded_objects <- readRDS("00data_AdultAbscess.rds")

# Assign back the objects to the environment, if needed
list2env(loaded_objects, .GlobalEnv)
```


### 1.2.2 Process Phyloseq
```{r}
process_phyloseq(ps_fs)


#Rename one problematic column
meta_df <- as.data.frame(sample_data(ps_fs))
colnames(meta_df)[colnames(meta_df) == "X.SampleID"] <- "SampleID"
sample_data(ps_fs) <- sample_data(meta_df)
sample_data(ps_f) <- sample_data(meta_df)

```


# Quick code for heatmap
```{r}
ps_genus <- tax_glom(ps_fs, taxrank = "Genus")

# Extract count matrix and convert to matrix
otu_mat <- as(otu_table(ps_genus), "matrix")

# Transpose if samples are rows (optional, depending on orientation)
if(taxa_are_rows(ps_genus)) {
  otu_mat <- t(otu_mat)
}

# Extract Genus names
taxa_names <- as.data.frame(tax_table(ps_genus))$Genus

# Plot heatmap
pheatmap(log(otu_mat + .0000001),
         scale = "none",        # keep raw counts
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         fontsize_row = 8,
         fontsize_col = 8,
         show_rownames = TRUE,
         show_colnames = TRUE)
```


# 2.0 Ordinations

# Ordination parameter sweep
```{r}
result_df <- compare_ordination_methods(
  phyloseq_obj = ps_f,
  rank_transformations = c("Phylum", "Genus"),          
  trans_types = c("identity", "clr", "log", "relab"),           
  dist_cal_types = c("euclidean", "bray", "jaccard"), 
  ord_calc_methods = c("NMDS", "PCA", "PCoA"),
  No_axes_to_correlated_with_LibSize = 2,
  variance_threshold = 0.80  # Set variance threshold 
)


result_df
```


## Organize results
```{r}
#Identify spearman_rho.* columns
rho_cols <- grep("^spearman_rho", colnames(result_df), value = TRUE)

#Compute absolute sum of all spearman_rho columns row-wise
df_with_rho_sum <- result_df %>%
  mutate(rho_abs_sum = apply(select(., all_of(rho_cols)), 1, function(x) sum(abs(x), na.rm = TRUE)))

#Group, split, sort each group, and recombine
grouped_list <- df_with_rho_sum %>%
  group_by(rank_transformation, ord_calc_method) %>%
  group_split()

#Apply base R function to sort each group
sorted_list <- lapply(grouped_list, function(g) {
  g[order(-g$rho_abs_sum), ]
})

#Combine sorted groups back into one data frame
df_sorted <- bind_rows(sorted_list)

df_sorted <- df_sorted %>%
  group_by(rank_transformation, ord_calc_method) %>%
  mutate(rho_rank = row_number(rho_abs_sum)) %>%
  ungroup()

df_sorted %>% filter(rho_rank == 1 | rho_rank == 2 )
```


## NMDS Genus relab bray
```{r, fig.width=5, fig.height=3.5}
Genus_NMDS_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Genus",
  trans_type = "relab",
  ord_calc_method = "NMDS",
  dist_cal_type = "bray",  #only needed for PCoA, PCA uses euclidean by default
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Genus_NMDS_res$plot
```

```{r, fig.width=3, fig.height=3}
Genus_NMDS_res_scores_df <- as.data.frame(Genus_NMDS_res$scores)

Genus_NMDS_res_scores_df_sil <- analyze_type_clustering_on_pca(Genus_NMDS_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Genus_NMDS_res_scores_df_sil$sil_plot
```

## PCoA Genus relab bray
```{r, fig.width=5, fig.height=3.5}
Genus_PCoA_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Genus",
  trans_type = "relab",
  ord_calc_method = "PCoA",
  dist_cal_type = "bray", 
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Genus_PCoA_res$plot
```

```{r, fig.width=3, fig.height=3}
Genus_PCoA_res_scores_df <- as.data.frame(Genus_PCoA_res$scores)

Genus_PCoA_res_scores_df_sil <- analyze_type_clustering_on_pca(Genus_PCoA_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Genus_PCoA_res_scores_df_sil$sil_plot
```


## PCA Genus relab 
```{r, fig.width=5, fig.height=3.5}
Genus_PCA_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Genus",
  trans_type = "relab",
  ord_calc_method = "PCA",
  dist_cal_type = "euclidean",  #only needed for PCoA, PCA uses euclidean by default
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Genus_PCA_res$plot
```


```{r, fig.width=3, fig.height=3}
Genus_PCA_res_scores_df <- as.data.frame(Genus_PCA_res$scores)

Genus_PCA_res_scores_df_sil <- analyze_type_clustering_on_pca(Genus_PCA_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Genus_PCA_res_scores_df_sil$sil_plot
```



## NMDS Phylum relab bray
```{r, fig.width=5, fig.height=3.5}
Phylum_NMDS_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Phylum",
  trans_type = "relab",
  ord_calc_method = "NMDS",
  dist_cal_type = "bray",  #only needed for PCoA, PCA uses euclidean by default
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Phylum_NMDS_res$plot
```

```{r, fig.width=3, fig.height=3}
Phylum_NMDS_res_scores_df <- as.data.frame(Phylum_NMDS_res$scores)

Phylum_NMDS_res_scores_df_sil <- analyze_type_clustering_on_pca(Phylum_NMDS_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Phylum_NMDS_res_scores_df_sil$sil_plot
```

## PCoA Phylum relab jaccard
```{r, fig.width=5, fig.height=3.5}
Phylum_PCoA_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Phylum",
  trans_type = "relab",
  ord_calc_method = "PCoA",
  dist_cal_type = "bray", 
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Phylum_PCoA_res$plot
```

```{r, fig.width=3, fig.height=3}
Phylum_PCoA_res_scores_df <- as.data.frame(Phylum_PCoA_res$scores)

Phylum_PCoA_res_scores_df_sil <- analyze_type_clustering_on_pca(Phylum_PCoA_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Phylum_PCoA_res_scores_df_sil$sil_plot
```


## PCA Phylum relab 
```{r, fig.width=5, fig.height=3.5}
Phylum_PCA_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Phylum",
  trans_type = "relab",
  ord_calc_method = "PCA",
  dist_cal_type = "euclidean",  #only needed for PCoA, PCA uses euclidean by default
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Phylum_PCA_res$plot
```


```{r, fig.width=3, fig.height=3}
Phylum_PCA_res_scores_df <- as.data.frame(Phylum_PCA_res$scores)

Phylum_PCA_res_scores_df_sil <- analyze_type_clustering_on_pca(Phylum_PCA_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Phylum_PCA_res_scores_df_sil$sil_plot
```




# 3.0 Alpha diversity
```{r, fig.width=4, fig.height=4}
a_my_comparisons <- list( c("Abscess", "Plaque"))
symnum.args = list(cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), symbols = c("****", "***", "**", "*", "ns"))

p_Shannon <- plot_richness(ps_f, x="Type", measures="Shannon", color = "Type")+
  geom_boxplot(alpha=0.6)+ 
    theme_bw(base_size=14) +
  theme(legend.position="none",         
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12), 
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=12), 
        strip.background = element_rect(fill = "white", color = "white"), 
        strip.text = element_text(size = 14, face = "bold"))  +
  stat_compare_means(method = "wilcox.test", comparisons = a_my_comparisons, label = "p.format")+
  scale_color_manual(values = colors_all) +
  ylim(0, 6.2)



p_Observed <- plot_richness(ps_f, x="Type", measures="Observed", color = "Type")+
  geom_boxplot(alpha=0.6)+ 
    theme_bw(base_size=14) +
  theme(legend.position="none",         
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12), 
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=12), 
        strip.background = element_rect(fill = "white", color = "white"), 
        strip.text = element_text(size = 14, face = "bold"))  +
  stat_compare_means(method = "wilcox.test", comparisons = a_my_comparisons, label = "p.format")+
  scale_color_manual(values = colors_all) +
  ylim(0, 470)

plot <- ggarrange(p_Shannon, p_Observed, ncol=2, nrow=1, common.legend = TRUE, legend="bottom")
annotate_figure(plot, top = text_grob(paste0("Alpha Diversity"), color = "black", face = "bold", size = 14))
```




# 4.0 Taxa Plots

## 4.1 Average Bar Plots

### 4.1.1 Phylum 
```{r taxa_phylum, fig.height=4, fig.width=4}

top_df <- ps_fs_t_n %>%
  tax_glom(., "Phylum") %>%
  get_top_taxa(., 10, relative = TRUE, discard_other = TRUE) %>% psmelt(.)
top_df$Phylum <- gsub("p__","",as.character(top_df$Phylum))

top_df %>%
  dplyr::group_by(Type,Phylum)%>%
  dplyr::summarise(Average_Abundance = mean(Abundance))%>%
  ggbarplot(x= "Type", y="Average_Abundance",
            fill = "Phylum",
            panel.labs.font = list(size=12),
            panel.labs.background = list(color = NULL, fill = "white"),
            font.tickslab = 14,
            font.legend = c(10),
            palette = core_colors, 
            strip.position = "top",
            color = "black",
            x.text.angle = 45, 
            y.text.angle = 0, 
            font.x = 14,
            font.y = 12,
            font.title = 16,
            rotate = FALSE,
            title = "Phlyum Level", 
            xlab = "",
            ylab = "Average Relative Abundance",
            ggtheme = theme_bw()) + 
  font("title", face="bold") +
  theme(plot.title=element_text(hjust=0.5), plot.subtitle=element_text(hjust=0.5))
```

### 4.1.2 Genus 
```{r taxa_genus, fig.height=4, fig.width=3}

top_df <- ps_fs_t_n %>%
  tax_glom(., "Genus") %>%
  get_top_taxa(., 10, relative = TRUE, discard_other = TRUE) %>% psmelt(.)
top_df$Genus <- gsub("g__","",as.character(top_df$Genus))

top_df %>%
  dplyr::group_by(Type,Genus)%>%
  dplyr::summarise(Average_Abundance = mean(Abundance))%>%
  ggbarplot(x= "Type", y="Average_Abundance",
            fill = "Genus",
            panel.labs.font = list(size=12),
            panel.labs.background = list(color = NULL, fill = "white"),
            font.tickslab = 14,
            font.legend = c(10),
           # palette = core_colors, 
            strip.position = "top",
            color = "black",
            x.text.angle = 45, 
            y.text.angle = 0, 
            font.x = 14,
            font.y = 12,
            font.title = 16,
            rotate = FALSE,
            title = "Top 10 Genera", 
            xlab = "",
            ylab = "Average Relative Abundance",
            ggtheme = theme_bw()) + 
  font("title", face="bold") +
  theme(plot.title=element_text(hjust=0.5), plot.subtitle=element_text(hjust=0.5), legend.text = element_text(face = "italic"))
```

### 4.1.3 Microshades Genus MB
```{r, fig.width=8, fig.height=4}
# Use microshades function prep_mdf to agglomerate, normalize, and melt the phyloseq object
mdf_prep <- prep_mdf(ps_fs)

mdf_prep <- mdf_prep %>%
  mutate(Kingdom = gsub("k__", "", Kingdom)) %>%
  mutate(Phylum = gsub("p__", "", Phylum)) %>%
  mutate(Class = gsub("c__", "", Class)) %>%
  mutate(Order = gsub("o__", "", Order)) %>%
  mutate(Family = gsub("f__", "", Family)) %>%
  mutate(Genus = gsub("g__", "", Genus)) %>%
  mutate(Species = gsub("s__", "", Species)) 

print(unique(mdf_prep$Phylum))

# Create a color object for the specified data
color_obj <- create_color_dfs(mdf_prep, 
            selected_groups = c("Proteobacteria", "Actinobacteria","Bacteroidetes", "Fusobacteria",
                                "Spirochaetes"),
    cvd = FALSE, top_orientation=TRUE)

# Extract
mdf <- color_obj$mdf
cdf <- color_obj$cdf

plot_1 <- plot_microshades(mdf, cdf)

plot_1 + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme_bw() +
  theme(legend.key.size = unit(0.2, "cm"), text=element_text(size=10), ) +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~Type, scales= "free_x") 
```



### 4.1.4 Species 
```{r species}
top_df <- ps_fs_t_n %>%
  tax_glom(., "Species") %>%
  get_top_taxa(., 10, relative = TRUE, discard_other = TRUE) %>% psmelt(.)
top_df$Species <- gsub("s__","",as.character(top_df$Species))

top_df %>%
  dplyr::group_by(Type,Species)%>%
  dplyr::summarise(Average_Abundance = mean(Abundance))%>%
  ggbarplot(x= "Type", y="Average_Abundance",
            fill = "Species",
            panel.labs.font = list(size=12),
            panel.labs.background = list(color = NULL, fill = "white"),
            font.tickslab = 10,
            font.legend = c(10),
            palette = pal20, 
            strip.position = "top",
            color = "black",
            x.text.angle = 90, 
            y.text.angle = 0, 
            font.x = 12,
            font.y = 12,
            font.title = 16,
            rotate = TRUE,
            title = "", 
            subtitle = "Species level",
            xlab = "Condition",
            ylab = "Average Abundance",
            ggtheme = theme_bw()) + 
  font("title", face="bold") +
  theme(plot.title=element_text(hjust=0.5), plot.subtitle=element_text(hjust=0.5))
```


## 4.2 Per taxa plots

### 4.2.0 Phylum TSS Dot
```{r, fig.width=15, fig.height=15}
plot_all_taxa(ps_fs, "TSS", "Type", "Phylum", "dot", colors_all)
```


### 4.2.1 Genus TSS Dot
```{r, fig.width=15, fig.height=50}
plot_all_taxa(ps_fs, "TSS", "Type", "Genus", "dot", colors_all)
```


### 4.2.1 Genus CSS Dot
```{r, fig.width=15, fig.height=50}
plot_all_taxa(ps_fs, "CLR", "Type", "Genus", "dot", colors_all)
```


### 4.2.2 Selected Species Plot
```{r, fig.width=5, fig.height=4}
source("MK_Microbiome_Functions.R")

plot_all_taxa(ps_obj = ps_fs,
              transformation = "TSS",
              group = "Type",
              taxa_level = "Species",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("g__Fusobacterium_s__nucleatum", "g__Fusobacterium_s__periodonticum"))
```



# Differential Abundance
## 6.1 ANCOMBC2

### 6.1.1 Phylum
```{r, fig.height=7, fig.width=10}
runANCOM(ps_fs, 
         tax_level = "Phylum", 
         group = "Type", 
         name_of_saved_results = "saved_analysis_files/ancombc_results_phy", 
         plot_type = "dot",
         Log2FC_cutoff = 0.5,
         plot_heights = c(1,2), 
         plot_colors = colors_all)
```

### 6.1.2 Genus
```{r, fig.height=30, fig.width=10}
runANCOM(ps_fs, 
         tax_level = "Genus", 
         group = "Type", 
         name_of_saved_results = "saved_analysis_files/ancombc_results_gen", 
         plot_type = "dot",
         plot_heights = c(1, 8), 
         plot_colors = colors_all )
```

### 6.1.3 Species
```{r, fig.height=45, fig.width=10}
runANCOM(ps_fs, 
         tax_level = "Species", 
         group = "Type", 
         name_of_saved_results = "saved_analysis_files/ancombc_results_sp", 
         plot_type = "dot",
         #Log2FC_cutoff = 0.5,
         plot_heights = c(1, 8),
         plot_colors = colors_all )
```


## 6.2 Maaslin2

### 6.2.1 Iterate Parameters
```{r, eval=FALSE, fig.height=8, fig.width=10}
resolutions <- c("Species", "Genus", "Phylum")
iterate_maaslin2(ps_obj = ps_fs, 
                 iterative_methods = iterative_methods,
                 resolutions = resolutions,
                 group = "Type",
                 qval_threshold = .25, 
                 colors_all = maaslin2_colors, 
                 percentage = TRUE)
```

### 6.2.2 Run Maaslin2

#### Phylum
```{r, fig.width=10, fig.height=30}
run_Maaslin2(ps_obj = ps_fs, 
                                 taxa_level = "Phylum", 
                                 group = "Type",
                                 analysis_method = "LM", 
                                 normalization = "CSS",
                                 transform = "LOG", 
                                 plot_colors = colors_all,
                                 plot_type = "dot", 
                                 qval_threshold= 0.25, 
                                 plot_heights = c(1, 9))
```


#### Genus
```{r, fig.width=10, fig.height=65}
run_Maaslin2(ps_obj = ps_fs, 
                                 taxa_level = "Genus", 
                                 group = "Type",
                                 analysis_method = "LM", 
                                 normalization = "CSS",
                                 transform = "LOG", 
                                 plot_colors = colors_all,
                                 plot_type = "dot", 
                                 qval_threshold= 0.25, 
                                 plot_heights = c(1, 9))
```

#### Species
```{r, fig.width=15, fig.height=15}
run_Maaslin2(ps_obj = ps_fs, 
                                 taxa_level = "Species", 
                                 group = "Type",
                                 analysis_method = "LM", 
                                 normalization = "CSS",
                                 transform = "LOG", 
                                 plot_colors = colors_all,
                                 plot_type = "dot", 
                                 qval_threshold= 0.25, 
                                 plot_heights = c(1, 9))
```

## 6.3 AlDEX2

### 6.3.1 Iterate Parameters
https://www.nature.com/articles/s41467-022-28034-z

"...performed CLR transformation of each realization, and then performed Wilcoxon tests on the transformed realizations. The function then returned the expected Benjamini-Hochberg (BH) FDR-corrected p-value for each feature based on the results the different across Monte Carlo samples."

ALDEx2 uses by default the centred log-ratio (clr) transformation which is scale invariant

```{r, eval=FALSE,  fig.width=20, fig.height=12}
resolutions <- c("Species", "Genus", "Phylum")

 
iterate_aldex2(ps_obj = ps_fs, 
                 iterative_methods = iterative_methods,
                 resolutions = resolutions,
                 group = "Type",
                 percentage = FALSE)

aldex2_iterative_summary_table %>%
  filter(resolution == "Genus") %>%
  filter(paired == FALSE) %>%
  filter(paired == FALSE) %>%
  filter(denom == "all") %>%
  filter(transform == "log10") %>%
  dplyr::arrange(desc(significant_features)) %>%
  filter(significant_features > 0)

```

### 6.3.2 Run Aldex2

#### Phylum
```{r, fig.width=10, fig.height=3}
run_aldex2(ps_fs, 
                  "Type", 
                  "Phylum",
                  method = "wilcox.test",
                  transform = "log10",
                  normalization = "CSS", 
                 plot_colors = colors_all)
```


#### Genus
```{r, fig.width=7, fig.height=5}
run_aldex2(ps_fs, 
                  "Type", 
                  "Genus",
                  method = "wilcox.test",
                  transform = "log10",
                  normalization = "CSS", 
                 plot_colors = colors_all)
```

#### Species
```{r, fig.width=12}
run_aldex2(ps_fs, 
                  "Type", 
                  "Species",
                  method = "wilcox.test",
                  transform = "log10",
                  normalization = "CSS", 
            plot_colors = colors_all)
```


## 6.4 Combine results


### Phylum
```{r, fig.height=10, fig.width=7}
DA_results_df_Phylum <- combine_DA(maaslin2_results_Phylum, ancombc2_results_df_Phylum, aldex2_res_Phylum, "Type", "Phylum", c(1,4), plot_colors=colors_all)
```


### Genus
```{r, fig.height=6, fig.width=5}
DA_results_df_genus <- combine_DA(maaslin2_results_Genus, ancombc2_results_df_Genus, aldex2_res_Genus, "Type", "Genus", c(1,3), plot_colors=colors_all)
```

### Species
```{r, fig.height=10, fig.width=7}
DA_results_df <- combine_DA(maaslin2_results_Species, ancombc2_results_df_Species, aldex2_res_Species, "Type", "Species", c(1,4), plot_colors=colors_all)
```

```{r, fig.width=4, fig.height=3}
plot_all_taxa(ps_obj = ps_fs,
              transformation = "TSS",
              group = "Type",
              taxa_level = "Species",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("g__Fusobacterium_s__nucleatum", "g__Fusobacterium_s__periodonticum"))
```


# 7.0 Random Forest
```{r}
library(randomForest)
library(caret)
set.seed(1234)

df <- ps_fs_genus_csv %>% column_to_rownames(var="Sample")
k <- 4

# Create stratified folds (list of test indices for each fold)
folds <- createFolds(df$Type, k = k, returnTrain = TRUE)

# View class distribution in each fold
for (i in seq_along(folds)) {
  cat("Fold", i)
  print(table(df[folds[[i]], "Type"]))
  cat("\n")
}

# Set up trainControl with custom indices
fitControl <- trainControl(
  method = "cv",
  number = k,
  index = folds,  # custom folds for training
  classProbs = TRUE,
  savePredictions = TRUE,
  summaryFunction = twoClassSummary
)

# Train the random forest model
set.seed(123)
model <- train(
  Type ~ .,
  data = df,
  method = "rf",
  trControl = fitControl,
  metric = "ROC"
)

```

```{r, fig.width=4, fig.height=3}
#Plot ROC Curve for each fold of the model
for_lift <- data.frame(Class = model$pred$obs, rf = model$pred$Plaque, resample = model$pred$Resample)
lift_df <-  data.frame()

for (fold in unique(for_lift$resample)) {
  fold_df <- dplyr::filter(for_lift, resample == fold)
  lift_obj_data <- lift(Class ~ rf, data = fold_df, class = "Plaque")$data
  lift_obj_data$fold = fold
  lift_df = rbind(lift_df, lift_obj_data)
}

lift_obj <- lift(Class ~ rf, data = for_lift, class = "Plaque")

ggplot(lift_df) +
  geom_line(aes(x = 1 - Sp, y = Sn, color = fold)) +
  scale_color_manual(
    name = "Fold",
    values = c("#FF495C", "#03CEA4", "#BBCC00", "#3399FF", "#5F00BA")
  ) +
  theme_bw()
```



### Print AUC Values
```{r}
AUC_score <-ddply(model$pred, "Resample", summarise,
      accuracy = Accuracy(pred, obs))

AUC_score %>% kbl() %>%
  kable_minimal()
```


### Print out top features
```{r}
# Extract variable importance
importance <- varImp(model, scale = FALSE)

# Convert to data frame for easier manipulation
importance_df <- as.data.frame(importance$importance)
importance_df$Feature <- rownames(importance_df)

# Print top 10 most important features
top_features <- importance_df %>%
  arrange(dplyr::desc(Overall)) %>%
  filter(Overall > 0 )

kable(top_features, format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                position = "center")
```


# 8.0 Core microbiome
## Core Microbiome
```{r}
library(microbiome)

core_ps <- ps_fs
core_ps_n <- transform_sample_counts(core_ps, function(x) (x) / sum(x))

# Agglomerate to the genus level
core_ps_genus <- tax_glom(core_ps, taxrank = "Genus")

# Initialize an empty dataframe to store results
results <- data.frame(prevalence = numeric(), percent_OTUs = numeric())

# Get the total number of OTUs at the species level
total_OTUs <- length(taxa_names(core_ps_genus))

# Loop through prevalence values from 0.1 to 1
for (p in seq(0.1, 1, by = 0.1)) {
  core.taxa <- core_members(core_ps_genus, detection = 0.00001, prevalence = p)
  num_OTUs <- length(core.taxa)
  percent_OTUs <- (num_OTUs / total_OTUs) * 100
  
  # Append the results to the dataframe
  results <- rbind(results, data.frame(prevalence = p*100, percent_OTUs = percent_OTUs))
}

# Create a line graph using ggplot2
ggplot(results, aes(x = prevalence, y = percent_OTUs)) +
  geom_line() +
  geom_point() +
  labs(title = "Percent of Total Taxa Detectioned",
       x = "Percent of Samples Containing Taxa",
       y = "Percent of Total Taxa") +
  theme_minimal()

```

## Core microbiome heatmap
```{r, fig.height=5, fig.width=5}

# Get the total list of species
genus_list <- taxa_names(core_ps_genus) 

# Initialize an empty list to store core genera at each prevalence level
core_genus_list <- list()

# Loop through prevalence values 
for (p in seq(0.5, 1, by = 0.05)) {
  core.taxa <- core_members(core_ps_genus, detection = 0.00001, prevalence = p)
  core_genus_list[[as.character(p)]] <- core.taxa
}

# Create a binary matrix indicating the presence (1) or absence (0) of genera in the core microbiome
core_matrix <- matrix(0, nrow = length(genus_list), ncol = length(core_genus_list),
                      dimnames = list(genus_list, names(core_genus_list)))

for (p in names(core_genus_list)) {
  core_matrix[core_genus_list[[p]], p] <- 1
}

# Remove rows and columns that sum to zero
core_matrix <- core_matrix[rowSums(core_matrix) > 0, colSums(core_matrix) > 0]

# Replace OTU names with genus names in the tax_table
tax_table_genus <- as.data.frame(tax_table(core_ps_genus))
genus_names <- tax_table_genus[rownames(core_matrix), "Genus"]

# Create a new matrix with genus names as rownames
core_matrix_genus <- core_matrix
genus_names <- gsub("g__", "", genus_names)

rownames(core_matrix_genus) <- genus_names


# Sort the matrix by row sums
row_sums <- rowSums(core_matrix_genus)
core_matrix_genus <- core_matrix_genus[order(row_sums, decreasing = TRUE), ]



# Convert the data frame to row-wise format while keeping row names
core_df_genus <- core_matrix_genus %>% as.data.frame() %>% rownames_to_column(var="Genera") %>%
  rowwise() %>%
  mutate(row_total = sum(c_across(starts_with("0.")))) %>%
  arrange(dplyr::desc(row_total)) %>%
  ungroup() %>%
  mutate(Genera = factor(Genera, levels = unique(Genera))) %>%
  dplyr::select(-row_total) %>% 
  pivot_longer(cols = -Genera,
               names_to = "prevalence",
               values_to = "value") 

# Pivoting the data longer and plotting
plot <- ggplot(core_df_genus, aes(x = prevalence, y = Genera, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "dodgerblue2", name = "Prevalence") +
  labs(title = "Core Genera at Different Prevalence Levels",
       x = "Prevalence Level",
       y = "Genera") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8, face = "italic"),  # <-- italic here
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )

plot

```

# 9.0 Topic modeling

## Iterate through scaling factors and topic numbers to choose the ideal combination
```{r}
iterate_res <- iterate_scaling_factors(ps_fs, 
                        "Genus",
                        scale_factors = c(10, 100, 1000, 10000), 
                        k_values= c(2, 3, 4, 5, 6), 
                        auto_chose_k = FALSE, 
                        type_column = "Type", 
                        method = "Gibbs")

iterate_res$results %>%
  mutate(
    sil_sum = sil_median_Abscess + sil_median_Plaque,
    sil_rank = min_rank(dplyr::desc(sil_sum))) %>% # Rank from highest to lowest
    arrange(sil_rank) %>% kable() 

# %>%
#     kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
#                 full_width = F,
#                 position = "center")
```

## Prep and scale input
```{r}
results_scaling <- prep_data_scale(ps_fs, "Genus", 1000, "Type")
meta_data <- results_scaling$meta_data
counts_data <- results_scaling$counts_data
```

## Use FindTopicsNumber() 
Note: There is no alpha value here
```{r}
topics <- seq(from = 2, to = 10, by = 1)
res <- RunFindTopicsNumber(counts_data, topics, "Gibbs")
res$metrics_df
```


## LDA: Create the model
You can use the info from the optimum topic numbers printed out in the above function, or adjust as needed.
```{r}
result <- create_topic_model(counts_data, 2, TRUE, "Gibbs")
model <- result$lda_model
```

## Plot beta
```{r, fig.height=4, fig.width=10}
plot_beta(result, 10, result$b_df)
```

## Heatmap of gamma scores 
```{r, fig.height=2, fig.width=9}
heatmap_gamma(result, "Type", result$g_df)
```

## UMAP of gamma scores
```{r fig.height=6, fig.width=6}
library(umap)
type_colors <- c("Abscess" = "red", "Plaque" = "blue")
plot_gamma_umap(results, "Type", type_colors, result$g_df)
```

##Membership of topic by Type
```{r, fig.height=3.5, fig.width=4}
topic_membership(result, "Type", type_colors, result$g_df)
```


## Heatmap of rel abundance in original data of top taxa
```{r, fig.height=3, fig.width=7}
relab_heatmap(result, ps_fs, "Genus", "Type", topic_no=1, n_top_words=10, result$b_df)

relab_heatmap(result, ps_fs, "Genus", "Type", topic_no=2, n_top_words=10, result$b_df)
```



```{r}
#This is kind of a weird heatmap, I'm not sure it's actualy usful. 

posterior_results <- posterior(model)
topics_terms <- posterior_results$terms 

# Blank out most term names
term_names <- colnames(topics_terms)
term_labels <- ifelse((seq_along(term_names) %% 5) == 0, term_names, "")

# Apply these as column names
colnames(topics_terms) <- term_labels

pheatmap(log(topics_terms), 
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         scale = "none", 
         show_rownames = TRUE, 
         show_colnames = TRUE)

```






