---
title: "01 Cohort Analysis"
author: "Madeline Krieger"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: "cosmo"
    code_folding: "show"
params:
    output_suffix: NULL

---

```{r setup, include=FALSE}
library(config)

#CHANGE THIS CHUNK based on if you want to run Pediatric (default) or Adult Abscess (adult)
cfg <- config::get(config = "default")

knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      error = FALSE,
                      fig.path = cfg$figs,
                      fig.keep = "all",
                      dev = c("png"),
                      cache = FALSE,
                      # Do not use lazy loading which fails for large objects
                      cache.lazy = FALSE)
```


# 1.0 Load and Define

## 1.1 Load packages and presets
### 1.1.1 Packages
```{r 1.1_load-packages}
library("pacman")
pacman::p_load("rmarkdown", "tidyverse", "phyloseq", "knitr","kableExtra", "here", "plyr", "ggpubr", "microViz", "microbiome", "pheatmap", "vegan", "reshape2", "magrittr", "microshades", "pheatmap","vegan", "data.table", "Polychrome", "fantaxtic","cetcolor", "cowplot", "MicrobiomeStat", "randomForest", "caret", "mlbench", "MLmetrics", "mia", "here", "patchwork", "digest", "ANCOMBC", "Maaslin2", "microbiomeMarker", "fantaxtic", "topicmodels", "ldatuning", "umap", "tidytext", "cluster", "randomForest", "caret", "microbiome", "umap", "viridis")

here::i_am("01_cohort_analysis.Rmd")
``` 


### 1.1.2 My packages
```{r}
source("MK_Microbiome_Functions.R")
```



### 1.1.2 Colors and presents
```{r}
colors_all= c("Abscess" = "#FF495C", "Plaque"="#083D77")

core_colors = c("#FAAA00", "#3399FF","#F76F8E","#083D77","#B8D4E3", "#FF495C","#477071", '#03CEA4',  "#5F00BA", "#BDAC9E", "white", "#FFD900")

maaslin2_colors= c("CLR_LOG" = "#2c45b5", "CLR_NONE"="#86B8FD", 
                   "CSS_LOG" = "#FF495C", "CSS_NONE"="#F2929A",
                   "TMM_LOG" = "#F4B701", "TMM_NONE"="#FEE59A",
                   "TSS_LOG" = "#8FD694", "TSS_NONE"="#D2EED4",
                   "TSS_LOGIT" = "#354F52", "TSS_AST"="#CEDDDF")

pal20 <- createPalette(22, c("#F76F8E","#03CEA4", "#083D77"), range = c(30, 80))
pal20 <- unname(as.list(pal20))

resolution <- c("Phylum", "Class", "Order", "Family", "Genus", "Species")

set.seed(1234)
date = Sys.Date()
```


## 1.2 Load data and process

### 1.2.1 Load Data from 00
```{r}
# Load the saved Rds file
loaded_objects <- readRDS(cfg$rds)

# Assign back the objects to the environment, if needed
list2env(loaded_objects, .GlobalEnv)
```


### 1.2.2 Process Phyloseq
```{r}
process_phyloseq(ps_fs)
```


# 2.0 Ordinations

# Ordination parameter sweep
```{r}
result_df <- compare_ordination_methods(
  phyloseq_obj = ps_f,
  rank_transformations = c("Phylum", "Genus"),          
  trans_types = c("identity", "clr", "log", "relab"),           
  dist_cal_types = c("euclidean", "bray", "jaccard"), 
  ord_calc_methods = c("NMDS", "PCA", "PCoA"),
  No_axes_to_correlated_with_LibSize = 2,
  variance_threshold = 0.80  # Set variance threshold 
)

result_df <- result_df$results_df
```


## Organize results
```{r}
#Identify spearman_rho.* columns
rho_cols <- grep("^spearman_rho", colnames(result_df), value = TRUE)

#Compute absolute sum of all spearman_rho columns row-wise
df_with_rho_sum <- result_df %>%
  mutate(rho_abs_sum = apply(select(., all_of(rho_cols)), 1, function(x) sum(abs(x), na.rm = TRUE)))

#Group, split, sort each group, and recombine
grouped_list <- df_with_rho_sum %>%
  group_by(rank_transformation, ord_calc_method) %>%
  group_split()

#Apply base R function to sort each group
sorted_list <- lapply(grouped_list, function(g) {
  g[order(-g$rho_abs_sum), ]
})

#Combine sorted groups back into one data frame
df_sorted <- bind_rows(sorted_list)

df_sorted <- df_sorted %>%
  group_by(rank_transformation, ord_calc_method) %>%
  mutate(rho_rank = row_number(rho_abs_sum)) %>%
  ungroup()

df_sorted %>% filter(rho_rank == 1 | rho_rank == 2 ) %>% kable()
```


## NMDS Genus relab bray
```{r, fig.width=5, fig.height=3.5}
Genus_NMDS_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Genus",
  trans_type = "relab",
  ord_calc_method = "NMDS",
  dist_cal_type = "bray",  #only needed for PCoA, PCA uses euclidean by default
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Genus_NMDS_res$plot
```

```{r, fig.width=3, fig.height=3}
Genus_NMDS_res_scores_df <- as.data.frame(Genus_NMDS_res$scores)

Genus_NMDS_res_scores_df_sil <- analyze_type_clustering_on_pca(Genus_NMDS_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Genus_NMDS_res_scores_df_sil$sil_plot
```

## PCoA Genus relab bray
```{r, fig.width=5, fig.height=3.5}
Genus_PCoA_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Genus",
  trans_type = "relab",
  ord_calc_method = "PCoA",
  dist_cal_type = "bray", 
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Genus_PCoA_res$plot
Genus_PCoA_res$lib_size_with_Axis1_corr
```

```{r, fig.width=3, fig.height=3}
Genus_PCoA_res_scores_df <- as.data.frame(Genus_PCoA_res$scores)

Genus_PCoA_res_scores_df_sil <- analyze_type_clustering_on_pca(Genus_PCoA_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Genus_PCoA_res_scores_df_sil$sil_plot
```


## PCA Genus relab 
```{r, fig.width=5, fig.height=3.5}
Genus_PCA_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Genus",
  trans_type = "relab",
  ord_calc_method = "PCA",
  dist_cal_type = "euclidean",  #only needed for PCoA, PCA uses euclidean by default
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Genus_PCA_res$plot
```


```{r, fig.width=3, fig.height=3}
Genus_PCA_res_scores_df <- as.data.frame(Genus_PCA_res$scores)

Genus_PCA_res_scores_df_sil <- analyze_type_clustering_on_pca(Genus_PCA_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Genus_PCA_res_scores_df_sil$sil_plot
```



## NMDS Phylum relab bray
```{r, fig.width=5, fig.height=3.5}
Phylum_NMDS_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Phylum",
  trans_type = "relab",
  ord_calc_method = "NMDS",
  dist_cal_type = "bray",  #only needed for PCoA, PCA uses euclidean by default
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Phylum_NMDS_res$plot
```

```{r, fig.width=3, fig.height=3}
Phylum_NMDS_res_scores_df <- as.data.frame(Phylum_NMDS_res$scores)

Phylum_NMDS_res_scores_df_sil <- analyze_type_clustering_on_pca(Phylum_NMDS_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Phylum_NMDS_res_scores_df_sil$sil_plot
```

## PCoA Phylum relab jaccard
```{r, fig.width=5, fig.height=3.5}
Phylum_PCoA_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Phylum",
  trans_type = "relab",
  ord_calc_method = "PCoA",
  dist_cal_type = "bray", 
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Phylum_PCoA_res$plot
Phylum_PCoA_res$lib_size_with_Axis1_corr
```

```{r, fig.width=3, fig.height=3}
Phylum_PCoA_res_scores_df <- as.data.frame(Phylum_PCoA_res$scores)

Phylum_PCoA_res_scores_df_sil <- analyze_type_clustering_on_pca(Phylum_PCoA_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Phylum_PCoA_res_scores_df_sil$sil_plot
```


## PCA Phylum relab 
```{r, fig.width=5, fig.height=3.5}
Phylum_PCA_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Phylum",
  trans_type = "relab",
  ord_calc_method = "PCA",
  dist_cal_type = "euclidean",  #only needed for PCoA, PCA uses euclidean by default
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Phylum_PCA_res$plot
```


```{r, fig.width=3, fig.height=3}
Phylum_PCA_res_scores_df <- as.data.frame(Phylum_PCA_res$scores)

Phylum_PCA_res_scores_df_sil <- analyze_type_clustering_on_pca(Phylum_PCA_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Phylum_PCA_res_scores_df_sil$sil_plot
```




# 3.0 Alpha diversity
```{r, fig.width=4, fig.height=4}
a_my_comparisons <- list( c("Abscess", "Plaque"))
symnum.args = list(cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), symbols = c("****", "***", "**", "*", "ns"))

p_Shannon <- plot_richness(ps_f, x="Type", measures="Shannon", color = "Type")+
  geom_boxplot(alpha=0.6)+ 
    theme_bw(base_size=14) +
  theme(legend.position="none",         
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12), 
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=12), 
        strip.background = element_rect(fill = "white", color = "white"), 
        strip.text = element_text(size = 14, face = "bold"))  +
  stat_compare_means(method = "wilcox.test", comparisons = a_my_comparisons, label = "p.format")+
  scale_color_manual(values = colors_all) +
  ylim(0, 6.2)



p_Observed <- plot_richness(ps_f, x="Type", measures="Observed", color = "Type")+
  geom_boxplot(alpha=0.6)+ 
    theme_bw(base_size=14) +
  theme(legend.position="none",         
        axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 12), 
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=12), 
        strip.background = element_rect(fill = "white", color = "white"), 
        strip.text = element_text(size = 14, face = "bold"))  +
  stat_compare_means(method = "wilcox.test", comparisons = a_my_comparisons, label = "p.format")+
  scale_color_manual(values = colors_all) +
  ylim(0, 470)

plot <- ggarrange(p_Shannon, p_Observed, ncol=2, nrow=1, common.legend = TRUE, legend="bottom")
annotate_figure(plot, top = text_grob(paste0("Alpha Diversity"), color = "black", face = "bold", size = 14))
```




# 4.0 Taxa Plots

## 4.1 Average Bar Plots

### 4.1.1 Phylum 
```{r taxa_phylum, fig.height=4, fig.width=4}

top_df <- ps_fs_t_n %>%
  tax_glom(., "Phylum") %>%
  get_top_taxa(., 10, relative = TRUE, discard_other = TRUE) %>% psmelt(.)
top_df$Phylum <- gsub("p__","",as.character(top_df$Phylum))

top_df %>%
  dplyr::group_by(Type,Phylum)%>%
  dplyr::summarise(Average_Abundance = mean(Abundance))%>%
  ggbarplot(x= "Type", y="Average_Abundance",
            fill = "Phylum",
            panel.labs.font = list(size=12),
            panel.labs.background = list(color = NULL, fill = "white"),
            font.tickslab = 14,
            font.legend = c(10),
            palette = core_colors, 
            strip.position = "top",
            color = "black",
            x.text.angle = 45, 
            y.text.angle = 0, 
            font.x = 14,
            font.y = 12,
            font.title = 16,
            rotate = FALSE,
            title = "Phlyum Level", 
            xlab = "",
            ylab = "Average Relative Abundance",
            ggtheme = theme_bw()) + 
  font("title", face="bold") +
  theme(plot.title=element_text(hjust=0.5), plot.subtitle=element_text(hjust=0.5))
```

### 4.1.2 Genus 
```{r taxa_genus, fig.height=4, fig.width=3}

top_df <- ps_fs_t_n %>%
  tax_glom(., "Genus") %>%
  get_top_taxa(., 10, relative = TRUE, discard_other = TRUE) %>% psmelt(.)
top_df$Genus <- gsub("g__","",as.character(top_df$Genus))

top_df %>%
  dplyr::group_by(Type,Genus)%>%
  dplyr::summarise(Average_Abundance = mean(Abundance))%>%
  ggbarplot(x= "Type", y="Average_Abundance",
            fill = "Genus",
            panel.labs.font = list(size=12),
            panel.labs.background = list(color = NULL, fill = "white"),
            font.tickslab = 14,
            font.legend = c(10),
            palette = core_colors, 
            strip.position = "top",
            color = "black",
            x.text.angle = 45, 
            y.text.angle = 0, 
            font.x = 14,
            font.y = 12,
            font.title = 16,
            rotate = FALSE,
            title = "Top 10 Genera", 
            xlab = "",
            ylab = "Average Relative Abundance",
            ggtheme = theme_bw()) + 
  font("title", face="bold") +
  theme(plot.title=element_text(hjust=0.5), plot.subtitle=element_text(hjust=0.5), legend.text = element_text(face = "italic"))
```

### 4.1.3 Microshades Genus MB
```{r, fig.width=8, fig.height=4}
# Use microshades function prep_mdf to agglomerate, normalize, and melt the phyloseq object
mdf_prep <- prep_mdf(ps_fs)

mdf_prep <- mdf_prep %>%
  mutate(Kingdom = gsub("k__", "", Kingdom)) %>%
  mutate(Phylum = gsub("p__", "", Phylum)) %>%
  mutate(Class = gsub("c__", "", Class)) %>%
  mutate(Order = gsub("o__", "", Order)) %>%
  mutate(Family = gsub("f__", "", Family)) %>%
  mutate(Genus = gsub("g__", "", Genus)) %>%
  mutate(Species = gsub("s__", "", Species)) 

print(unique(mdf_prep$Phylum))

# Create a color object for the specified data
color_obj <- create_color_dfs(mdf_prep, 
            selected_groups = c("Proteobacteria", "Actinobacteria","Bacteroidetes", "Fusobacteria",
                                "Spirochaetes"),
    cvd = FALSE, top_orientation=TRUE)

# Extract
mdf <- color_obj$mdf
cdf <- color_obj$cdf

plot_1 <- plot_microshades(mdf, cdf)

plot_1 + scale_y_continuous(labels = scales::percent, expand = expansion(0)) +
  theme_bw() +
  theme(legend.key.size = unit(0.2, "cm"), text=element_text(size=10), ) +
  theme(axis.text.x = element_text(size= 6)) +
  facet_wrap(~Type, scales= "free_x") 
```



### 4.1.4 Species 
```{r species}
top_df <- ps_fs_t_n %>%
  tax_glom(., "Species") %>%
  get_top_taxa(., 10, relative = TRUE, discard_other = TRUE) %>% psmelt(.)
top_df$Species <- gsub("s__","",as.character(top_df$Species))

top_df %>%
  dplyr::group_by(Type,Species)%>%
  dplyr::summarise(Average_Abundance = mean(Abundance))%>%
  ggbarplot(x= "Type", y="Average_Abundance",
            fill = "Species",
            panel.labs.font = list(size=12),
            panel.labs.background = list(color = NULL, fill = "white"),
            font.tickslab = 10,
            font.legend = c(10),
            palette = pal20, 
            strip.position = "top",
            color = "black",
            x.text.angle = 90, 
            y.text.angle = 0, 
            font.x = 12,
            font.y = 12,
            font.title = 16,
            rotate = TRUE,
            title = "", 
            subtitle = "Species level",
            xlab = "Condition",
            ylab = "Average Abundance",
            ggtheme = theme_bw()) + 
  font("title", face="bold") +
  theme(plot.title=element_text(hjust=0.5), plot.subtitle=element_text(hjust=0.5))
```


## 4.2 Heatmap
### 4.2.1 Genus
```{r, fig.height=5, fig.width=6.5}
df <- ps_fs_genus_csv %>% select(-Type) %>% column_to_rownames(var="Sample") %>% as.matrix %>% t()
#df <- df[order(rowSums(df), decreasing = TRUE)[1:50], ] #select top 50

# Extract Type information from sample_data
type_annot <- data.frame(Type = ps_fs_genus_csv$Type, Sample = ps_fs_genus_csv$Sample) %>% column_to_rownames(var="Sample")
rownames(type_annot) <- colnames(df)  # Ensure match to colnames of otu_mat

# Manually define annotation colors
ann_colors <- list(Type = c(Plaque = "#083D77", Abscess = "#FF495C"))

my_colors <- viridis(100, option = "inferno")

# Plot heatmap with annotation
pheatmap(log(df),
         scale = "none",
         clustering_distance_rows = "correlation",
         clustering_distance_cols = "correlation",
         clustering_method = "complete",
         fontsize_row = 8,
         fontsize_col = 8,
         show_rownames = FALSE,
         show_colnames = FALSE,
         annotation_col = type_annot,
         annotation_colors = ann_colors,
         color = my_colors,
         border_color = NA)
```


# 5.0 Taxa trends

## Genus
```{r, fig.height=6, fig.width=4}

df_avg <- ps_fs_genus_csv %>%
  select(-Sample) %>%
  dplyr::group_by(Type) %>%
  dplyr::summarise(across(where(is.numeric), ~ median(.x, na.rm = TRUE)),
                   .groups = "drop") %>% as.data.frame()
dif_row <- df_avg %>%
  dplyr::summarise(across(where(is.numeric),
                   ~ .[Type == "Plaque"] - .[Type == "Abscess"])) %>%
  mutate(Type = "dif") %>%
  relocate(Type)

# Bind to the original averages
df_final <- bind_rows(df_avg, dif_row) %>% column_to_rownames(var="Type") %>% t() %>% as.data.frame() %>% dplyr::arrange(desc(abs(dif)))

# Sort, take top X by absolute difference
df_top <- df_final %>%
  dplyr::arrange(desc(abs(dif))) %>%
  slice_head(n = 20) %>%
  mutate(Genus = rownames(.))   # keep g__ names for plotting

# Remove "g__" from genus names
df_top <- df_top %>%
  mutate(Genus = str_remove(Genus, "^g__")) %>%
  mutate(Genus = str_replace_all(Genus, "_", " ")) %>%
  mutate(Genus = str_remove_all(Genus, "\\[|\\]"))

# Make Genus a factor ordered by dif
df_top <- df_top %>%
  mutate(Genus = factor(Genus, levels = Genus[order(abs(dif), decreasing = FALSE)]))

# Reshape to long format for ggplot
df_long <- df_top %>%
  select(Genus, Abscess, Plaque) %>%
  pivot_longer(cols = c(Abscess, Plaque),
               names_to = "Type",
               values_to = "Value")
# Bar plot
ggplot(df_long, aes(x = Genus, y = Value, fill = Type)) +
  geom_col(position = position_dodge(width = 0.9), width = 0.8) +
  geom_vline(xintercept = seq(1.5, length(unique(df_long$Genus)) - 0.5, 1),
             color = "black", size = 0.3) +
  labs(x = NULL, y = "Median Relative Abundance", fill = "Type") +
  theme_bw(base_size = 12) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "italic", size = 13, color="black"),
    axis.text.x = element_text(hjust = 1, angle=45, size = 13, color="black"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank() ) + 
  scale_fill_manual(values = colors_all) +
  coord_flip()

```

### Sample-specific patterns
```{r, fig.height=6, fig.width=1}

#Detatch plyr because it causes some issues
if ("package:plyr" %in% search()) {
  detach("package:plyr", unload = TRUE, character.only = TRUE)
}

taxa_pull <- df_long$Genus

#A lot of code to determine the ratio of Abscess to plaque (+1 for more in abscess and -1 for more in plaque)
data_pull <- ps_fs_genus_csv %>% 
  pivot_longer(cols=starts_with("g_"),, names_to = "Genus", values_to = "relab") %>%
  mutate(Genus = str_remove(Genus, "^g__")) %>%
  mutate(Genus = str_replace_all(Genus, "_", " ")) %>%
  mutate(Genus = str_remove_all(Genus, "\\[|\\]")) %>%
  dplyr::filter(Genus %in% taxa_pull) %>%
  mutate(Sample = str_replace_all(Sample, "-P", "")) %>%
  mutate(Sample = str_replace_all(Sample, "-A", "")) %>%
  pivot_wider(names_from = "Type", values_from = "relab") %>%
  mutate(Ratio = Abscess / Plaque, RatioFlag = if_else(Ratio > 1, 1, -1))

#Counting the number of samples for each taxa with plaque and abscess enrichment
df_summary <- data_pull %>%
  mutate(Genus = trimws(as.character(Genus))) %>%  # ensure clean text
  dplyr::group_by(Genus) %>%
  summarise(
    Abscess_enriched  = sum(RatioFlag == 1, na.rm = TRUE),
    Plaque_enriched = sum(RatioFlag == -1, na.rm = TRUE)
  ) 

df_summary <- df_summary[match(df_top$Genus, df_summary$Genus), ]
df_summary <- df_summary %>% column_to_rownames(var="Genus")

# Plot heatmap with annotation
pheatmap(df_summary,
         scale = "none",
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         clustering_method = "complete",
         fontsize_row = 8,
         fontsize_col = 12,
         fontsize_number = 10,
         show_rownames = FALSE,
         show_colnames = TRUE,
         border_color = "black",
         display_numbers = round(as.matrix(df_summary), 1),  # 1 decimal place
        color = colorRampPalette(c( "#AA7DCE", "white", "#28CCC7"))(100),  # Custom color palette
)
```


## Species
```{r, fig.height=6, fig.width=5}
df_avg <- ps_fs_species_csv %>%
  select(-Sample) %>%
  dplyr::group_by(Type) %>%
  dplyr::summarise(across(where(is.numeric), ~ median(.x, na.rm = TRUE)),
                   .groups = "drop") %>% as.data.frame()
dif_row <- df_avg %>%
  dplyr::summarise(across(where(is.numeric),
                   ~ .[Type == "Plaque"] - .[Type == "Abscess"])) %>%
  mutate(Type = "dif") %>%
  relocate(Type)

# Bind to the original averages
df_final <- bind_rows(df_avg, dif_row) %>% column_to_rownames(var="Type") %>% t() %>% as.data.frame() %>% dplyr::arrange(desc(abs(dif)))

# Sort, take top X by absolute difference
df_top <- df_final %>%
  rownames_to_column(var="Species") %>%
  filter(!grepl("Streptococcus", Species)) %>%
  filter(!grepl("Actinomyces", Species)) %>%
  dplyr::arrange(desc(abs(dif))) %>%
  slice_head(n = 20) 

# Remove "g__" from Species names
df_top <- df_top %>%
  mutate(Species = str_remove(Species, "^g__")) %>%
  mutate(Species = str_remove(Species, "s__")) %>%
  mutate(Species = str_replace_all(Species, "_", " ")) %>%
  mutate(Species = str_remove_all(Species, "\\[|\\]"))

# Make Species a factor ordered by dif
df_top <- df_top %>%
  mutate(Species = factor(Species, levels = Species[order(abs(dif), decreasing = FALSE)]))

# Reshape to long format for ggplot
df_long <- df_top %>%
  select(Species, Abscess, Plaque) %>%
  pivot_longer(cols = c(Abscess, Plaque),
               names_to = "Type",
               values_to = "Value")
# Bar plot
ggplot(df_long, aes(x = Species, y = Value, fill = Type)) +
  geom_col(position = position_dodge(width = 0.9), width = 0.8) +
  geom_vline(xintercept = seq(1.5, length(unique(df_long$Species)) - 0.5, 1),
             color = "black", size = 0.3) +
  labs(x = NULL, y = "Median Relative Abundance", fill = "Type") +
  theme_bw(base_size = 12) +
  theme(
    axis.text.y = element_text(hjust = 1, face = "italic", size = 13, color="black"),
    axis.text.x = element_text(hjust = 1, angle=45, size = 13, color="black"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank() ) + 
  scale_fill_manual(values = colors_all) +
  coord_flip()
```

### Sample-specific patterns
```{r, fig.height=6, fig.width=1}

taxa_pull <- df_long$Species

#A lot of code to determine the ratio of Abscess to plaque (+1 for more in abscess and -1 for more in plaque)
data_pull <- ps_fs_species_csv %>% 
  pivot_longer(cols=starts_with("g_"),, names_to = "Species", values_to = "relab") %>%
  mutate(Species = str_remove(Species, "^g__")) %>%
  mutate(Species = str_remove(Species, "s__")) %>%
  mutate(Species = str_replace_all(Species, "_", " ")) %>%
  mutate(Species = str_remove_all(Species, "\\[|\\]")) %>%
  dplyr::filter(Species %in% taxa_pull) %>%
  mutate(Sample = str_replace_all(Sample, "-P", "")) %>%
  mutate(Sample = str_replace_all(Sample, "-A", "")) %>%
  pivot_wider(names_from = "Type", values_from = "relab") %>%
  mutate(Ratio = Abscess / Plaque, RatioFlag = if_else(Ratio > 1, 1, -1))


#Counting the number of samples for each taxa with plaque and abscess enrichment
df_summary <- data_pull %>%
  mutate(Species = trimws(as.character(Species))) %>%  # ensure clean text
  dplyr::group_by(Species) %>%
  summarise(
    Abscess_enriched  = sum(RatioFlag == 1, na.rm = TRUE),
    Plaque_enriched = sum(RatioFlag == -1, na.rm = TRUE)
  ) 

df_summary <- df_summary[match(df_top$Species, df_summary$Species), ]
df_summary <- df_summary %>% column_to_rownames(var="Species")


# Plot heatmap with annotation
pheatmap(df_summary,
         scale = "none",
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         clustering_method = "complete",
         fontsize_row = 8,
         fontsize_col = 12,
         fontsize_number = 10,
         show_rownames = FALSE,
         show_colnames = TRUE,
         border_color = "black",
         display_numbers = round(as.matrix(df_summary), 1),  # 1 decimal place
        color = colorRampPalette(c( "#AA7DCE", "white", "#28CCC7"))(100),  # Custom color palette
)

```



# 6.0 Differential Abundance
## 6.1 ANCOMBC2

### 6.1.1 Phylum
```{r, fig.height=7, fig.width=10}
ANCOM_ph <- runANCOM(ps_fs, 
         tax_level = "Phylum", 
         group = "Type", 
         name_of_saved_results = "../saved_analysis_files/ancombc_results_phy", 
         plot_type = "dot",
         Log2FC_cutoff = 0.5,
         plot_colors = colors_all)
```

### 6.1.2 Genus
```{r, fig.height=10, fig.width=10}
ANCOM_genus <- runANCOM(ps_fs, 
         tax_level = "Genus", 
         group = "Type", 
         name_of_saved_results = "../saved_analysis_files/ancombc_results_gen", 
         plot_type = "dot",
         plot_colors = colors_all )
```

### 6.1.2 Species
```{r, fig.height=10, fig.width=10}
ANCOM_sp <- runANCOM(ps_fs, 
         tax_level = "Species", 
         group = "Type", 
         name_of_saved_results = "../saved_analysis_files/ancombc_results_sp", 
         plot_type = "dot",
         plot_colors = colors_all )
```


## 6.2 Maaslin2

### 6.2.1 Iterate Parameters
```{r, eval=FALSE, fig.height=8, fig.width=10}
resolutions <- c("Species", "Genus", "Phylum")
iterate_maaslin2(ps_obj = ps_fs, 
                 iterative_methods = iterative_methods,
                 resolutions = resolutions,
                 group = "Type",
                 qval_threshold = .25, 
                 percentage = TRUE, 
                 plot_colors = colors_all)
```

### 6.2.2 Run Maaslin2

#### Phylum
```{r, fig.width=10, fig.height=30}
MaaslinPhylum <- run_Maaslin2(ps_obj = ps_fs, 
                                 taxa_level = "Phylum", 
                                 group = "Type",
                                 analysis_method = "LM", 
                                 normalization = "CSS",
                                 transform = "LOG", 
                                 plot_colors = colors_all,
                                 plot_type = "dot", 
                                 qval_threshold= .25)
MaaslinPhylum$results
```


#### Genus
```{r, fig.width=10, fig.height=65}
MaaslinGenus <- run_Maaslin2(ps_obj = ps_fs, 
                                 taxa_level = "Genus", 
                                 group = "Type",
                                 analysis_method = "LM", 
                                 normalization = "CSS",
                                 transform = "LOG", 
                                 plot_colors = colors_all,
                                 plot_type = "dot", 
                                 qval_threshold= .25)

MaaslinGenus$results
```


#### Species
```{r, fig.width=10, fig.height=65}
Maaslinsp <- run_Maaslin2(ps_obj = ps_fs, 
                                 taxa_level = "Species", 
                                 group = "Type",
                                 analysis_method = "LM", 
                                 normalization = "CSS",
                                 transform = "LOG", 
                                 plot_colors = colors_all,
                                 plot_type = "dot", 
                                 qval_threshold= .25)

Maaslinsp$results
```


## 6.3 AlDEX2

### 6.3.1 Iterate Parameters
https://www.nature.com/articles/s41467-022-28034-z

"...performed CLR transformation of each realization, and then performed Wilcoxon tests on the transformed realizations. The function then returned the expected Benjamini-Hochberg (BH) FDR-corrected p-value for each feature based on the results the different across Monte Carlo samples."

ALDEx2 uses by default the centred log-ratio (clr) transformation which is scale invariant

```{r, eval=FALSE,  fig.width=20, fig.height=12}
resolutions <- c("Species", "Genus", "Phylum")
 
iterate_aldex2(ps_obj = ps_fs, 
                 iterative_methods = iterative_methods,
                 resolutions = resolutions,
                 group = "Type",
                 percentage = FALSE)

aldex2_iterative_summary_table %>%
  filter(resolution == "Genus") %>%
  filter(paired == FALSE) %>%
  filter(paired == FALSE) %>%
  filter(denom == "all") %>%
  filter(transform == "log10") %>%
  dplyr::arrange(desc(significant_features)) %>%
  filter(significant_features > 0)
```

### 6.3.2 Run Aldex2

#### Phylum
```{r, fig.width=10, fig.height=3}
ALDEX_ph <- run_aldex2(ps_fs, 
                  "Type", 
                  "Phylum",
                  method = "wilcox.test",
                  transform = "log10",
                  normalization = "CSS", 
                 plot_colors = colors_all)
ALDEX_ph$results
```


#### Genus
```{r, fig.width=7, fig.height=5}
ALDEX_genus <- run_aldex2(ps_fs, 
                  "Type", 
                  "Genus",
                  method = "wilcox.test",
                  transform = "log10",
                  normalization = "CSS", 
                 plot_colors = colors_all)
ALDEX_genus$result
```


#### Species
```{r, fig.width=7, fig.height=5}
ALDEX_sp <- run_aldex2(ps_fs, 
                  "Type", 
                  "Species",
                  method = "wilcox.test",
                  transform = "log10",
                  normalization = "CSS", 
                 plot_colors = colors_all)
ALDEX_sp$result
```


## 6.4 Combine results

### Phylum
```{r, fig.height=1.2, fig.width=4}

source("MK_Microbiome_Functions.R")

DA_results_df_Phylum <- combine_DA(maaslin2_results_Phylum, ancombc2_results_df_Phylum, aldex2_res_Phylum, "Type", "Phylum", plot_colors=colors_all)

DA_results_df_Phylum$bar_plot_high_confidence
```

```{r, fig.height=1.2, fig.width=4}
DA_results_df_Phylum$bar_plot_med_confidence
```

### Genus
```{r, fig.height=2, fig.width=4}
DA_results_df_genus <- combine_DA(maaslin2_results_Genus, ancombc2_results_df_Genus, aldex2_res_Genus, "Type", "Genus", plot_colors=colors_all)

DA_results_df_genus$bar_plot_high_confidence
```

```{r, fig.height=4.5, fig.width=5}
DA_results_df_genus$bar_plot_med_confidence
```



### Species
```{r, fig.height=2, fig.width=4}
DA_results_df_sp <- combine_DA(maaslin2_results_Species, ancombc2_results_df_Species, aldex2_res_Species, "Type", "Species", plot_colors=colors_all)

DA_results_df_sp$bar_plot_high_confidence
```

```{r, fig.height=4.5, fig.width=5}
DA_results_df_sp$bar_plot_med_confidence
```



## 6.5 Per taxa plots

### Selected Phylum Plot
```{r, fig.width=6, fig.height=2.8}
plot_all_taxa(ps_obj = ps_fs,
              transformation = "TSS",
              group = "Type",
              taxa_level = "Phylum",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("p__Proteobacteria", "p__Spirochaetes", "p__Saccharibacteria_(TM7)", "p__Synergistetes" ))
```

```{r, fig.width=2.5, fig.height=2.8}
plot_all_taxa(ps_obj = ps_fs,
              transformation = "TSS",
              group = "Type",
              taxa_level = "Phylum",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("p__Fusobacteria" ))
```
```{r, fig.width=2.5, fig.height=2.8}
plot_all_taxa(ps_obj = ps_fs,
              transformation = "CLR",
              group = "Type",
              taxa_level = "Phylum",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("p__Fusobacteria" ))
```
### Selected Genus Plot
```{r, fig.width=8, fig.height=3}
plot_all_taxa(ps_obj = ps_fs,
              transformation = "TSS",
              group = "Type",
              taxa_level = "Genus",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("g__Leptotrichia", "g__Corynebacterium", "g__Actinomyces", "g__Treponema", "g__Dialister", "g__Veillonellaceae_[G-1]"))
```

```{r, fig.width=8, fig.height=3}
plot_all_taxa(ps_obj = ps_fs,
              transformation = "TSS",
              group = "Type",
              taxa_level = "Genus",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("g__Rothia", "g__Neisseria", "g__Lautropia", "g__Lancefieldella", "g__Solobacterium", "g__Oribacterium"))
```

```{r, fig.width=2.5, fig.height=3}
plot_all_taxa(ps_obj = ps_fs,
              transformation = "CLR",
              group = "Type",
              taxa_level = "Genus",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("g__Fusobacterium"))

plot_all_taxa(ps_obj = ps_fs,
              transformation = "TSS",
              group = "Type",
              taxa_level = "Genus",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("g__Fusobacterium"))
```

### Selected Species Plot
```{r, fig.width=4, fig.height=3}
plot_all_taxa(ps_obj = ps_fs,
              transformation = "TSS",
              group = "Type",
              taxa_level = "Species",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("g__Fusobacterium_s__nucleatum", "g__Fusobacterium_s__periodonticum"))


plot_all_taxa(ps_obj = ps_fs,
              transformation = "CLR",
              group = "Type",
              taxa_level = "Species",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("g__Fusobacterium_s__nucleatum", "g__Fusobacterium_s__periodonticum"))
```



# 7.0 Random Forest
```{r}
set.seed(1234)

df <- ps_fs_genus_csv %>% column_to_rownames(var="Sample")
k <- 4

# Create stratified folds (list of test indices for each fold)
folds <- createFolds(df$Type, k = k, returnTrain = TRUE)

# View class distribution in each fold
for (i in seq_along(folds)) {
  cat("Fold", i)
  print(table(df[folds[[i]], "Type"]))
  cat("\n")
}

# Set up trainControl with custom indices
fitControl <- trainControl(
  method = "cv",
  number = k,
  index = folds,  # custom folds for training
  classProbs = TRUE,
  savePredictions = TRUE,
  summaryFunction = twoClassSummary
)

# Train the random forest model
set.seed(123)
model <- train(
  Type ~ .,
  data = df,
  method = "rf",
  trControl = fitControl,
  metric = "ROC"
)


```


### Rank features by fold and create heatmap
```{r, fig.width=3.8, fig.height=4}
rank_rf_feature_table <- function(df, folds, mtry = NULL, top_n = NULL, add_mean = TRUE) {
  
  if (!is.factor(df$Type)) df$Type <- factor(df$Type)
  p <- ncol(df) - 1L
  if (is.null(mtry)) mtry <- max(1L, floor(sqrt(p)))


  get_fold_ranks <- function(train_idx, fold_id) {
    fit <- caret::train(
      Type ~ .,
      data = df[train_idx, ],
      method = "rf",
      trControl = caret::trainControl(method = "none"),
      tuneGrid = data.frame(mtry = mtry),
      importance = TRUE
    )
    imp <- caret::varImp(fit)$importance %>%
      rownames_to_column("Feature")

    # collapse to a single score column named Score (usually "Overall" already)
    score_cols <- setdiff(names(imp), "Feature")
    if (length(score_cols) == 0) stop("No importance scores returned.")
    if (length(score_cols) == 1) {
      imp <- transmute(imp, Feature, Score = .data[[score_cols]])
    } else {
      imp <- imp %>% mutate(Score = rowMeans(across(all_of(score_cols)), na.rm = TRUE)) %>%
        select(Feature, Score)
    }

    # rank (1 = most important)
    imp <- imp[order(imp$Score, decreasing = TRUE), , drop = FALSE]
    imp$Rank <- seq_len(nrow(imp))

    # return Feature + FoldX rank column
    fold_col <- paste0("Fold", fold_id)
    out <- imp[, c("Feature", "Rank")]
    names(out)[2] <- fold_col
    out
  }

  fold_tables <- Map(function(idx, i) get_fold_ranks(idx, i), folds, seq_along(folds))

  # Join by Feature (NOT by rank) so each taxon is one row
  wide <- Reduce(function(x, y) merge(x, y, by = "Feature", all = TRUE), fold_tables)

  # Optionally keep only taxa that are in top_n in at least one fold
  if (!is.null(top_n)) {
    rank_cols <- grep("^Fold\\d+$", names(wide), value = TRUE)
    keep <- apply(wide[rank_cols], 1, function(v) any(!is.na(v) & v <= top_n))
    wide <- wide[keep, , drop = FALSE]
  }

  # Mean rank across folds (lower is better)
  if (add_mean) {
    rank_cols <- grep("^Fold\\d+$", names(wide), value = TRUE)
    wide$MeanRank <- rowMeans(wide[rank_cols], na.rm = TRUE)
    wide <- wide[order(wide$MeanRank), , drop = FALSE]
  }

  rownames(wide) <- NULL
  wide
}

set.seed(123)  # only if you want reproducible rankings
rank_table <- rank_rf_feature_table(df, folds, top_n = 10, add_mean = TRUE)

# Example: show the first few rows
rank_table %>% kable()


mat <- rank_table %>%
  select(-Feature, -MeanRank) %>%
  as.data.frame()
rownames(mat) <- rank_table$Feature
rownames(mat) <- sub("`","", rownames(mat)) #Need this here twice
rownames(mat) <- sub("`","", rownames(mat))

rownames(mat) <- sub("^g__","", rownames(mat))

labs <- rownames(mat)

# Make taxa labels italics
labs_expr <- parse(text = sprintf("italic('%s')",
                                  gsub("'", "\\\\'", labs, fixed = TRUE)))

pheatmap(log(mat),
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         labels_row = labs_expr,
         color = colorRampPalette(c("#7B8CDE", "white", "#C0E6DE"))(50),
        display_numbers = round(as.matrix(mat), 1),  # 1 decimal place
        fontsize_number = 10,
         main = "Feature Ranks per Fold")
```



```{r, fig.width=4, fig.height=3}
#Plot ROC Curve for each fold of the model
for_lift <- data.frame(Class = model$pred$obs, rf = model$pred$Plaque, resample = model$pred$Resample)
lift_df <-  data.frame()

for (fold in unique(for_lift$resample)) {
  fold_df <- dplyr::filter(for_lift, resample == fold)
  lift_obj_data <- lift(Class ~ rf, data = fold_df, class = "Plaque")$data
  lift_obj_data$fold = fold
  lift_df = rbind(lift_df, lift_obj_data)
}

lift_obj <- lift(Class ~ rf, data = for_lift, class = "Plaque")

ggplot(lift_df) +
  geom_line(aes(x = 1 - Sp, y = Sn, color = fold)) +
  scale_color_manual(
    name = "Fold",
    values = c("#DD1155", "#00CC99", "#F4B942", "#3399FF")
  ) +
  theme_bw()
```



### Print AUC Values
```{r}
library("plyr")
AUC_score <-ddply(model$pred, "Resample", summarise,
      accuracy = Accuracy(pred, obs))

AUC_score %>% kbl() %>%
  kable_minimal()


if ("package:plyr" %in% search()) {
  detach("package:plyr", unload = TRUE, character.only = TRUE)
}
```


### Print out top features
```{r}
# Extract variable importance
importance <- varImp(model, scale = FALSE)

# Convert to data frame for easier manipulation
importance_df <- as.data.frame(importance$importance)
importance_df$Feature <- rownames(importance_df)

# Print top 10 most important features
top_features <- importance_df %>%
  arrange(dplyr::desc(Overall)) %>%
  filter(Overall > 0 )

kable(top_features, format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                position = "center")
```


# 8.0 Core microbiome
## Core Microbiome
```{r}
core_ps <- ps_fs
core_ps_n <- transform_sample_counts(core_ps, function(x) (x) / sum(x))

# Agglomerate to the genus level
core_ps_genus <- tax_glom(core_ps, taxrank = "Genus")

# Initialize an empty dataframe to store results
results <- data.frame(prevalence = numeric(), percent_OTUs = numeric())

# Get the total number of OTUs at the species level
total_OTUs <- length(taxa_names(core_ps_genus))

# Loop through prevalence values from 0.1 to 1
for (p in seq(0.1, 1, by = 0.1)) {
  core.taxa <- core_members(core_ps_genus, detection = 0.00001, prevalence = p)
  num_OTUs <- length(core.taxa)
  percent_OTUs <- (num_OTUs / total_OTUs) * 100
  
  # Append the results to the dataframe
  results <- rbind(results, data.frame(prevalence = p*100, percent_OTUs = percent_OTUs))
}

# Create a line graph using ggplot2
ggplot(results, aes(x = prevalence, y = percent_OTUs)) +
  geom_line() +
  geom_point() +
  labs(title = "Percent of Total Taxa Detectioned",
       x = "Percent of Samples Containing Taxa",
       y = "Percent of Total Taxa") +
  theme_minimal()

```

## Core microbiome heatmap
```{r, fig.height=5, fig.width=5}

# Get the total list of species
genus_list <- taxa_names(core_ps_genus) 

# Initialize an empty list to store core genera at each prevalence level
core_genus_list <- list()

# Loop through prevalence values 
for (p in seq(0.5, 1, by = 0.05)) {
  core.taxa <- core_members(core_ps_genus, detection = 1, prevalence = p)
  core_genus_list[[as.character(p)]] <- core.taxa
}

# Create a binary matrix indicating the presence (1) or absence (0) of genera in the core microbiome
core_matrix <- matrix(0, nrow = length(genus_list), ncol = length(core_genus_list),
                      dimnames = list(genus_list, names(core_genus_list)))

for (p in names(core_genus_list)) {
  core_matrix[core_genus_list[[p]], p] <- 1
}

# Remove rows and columns that sum to zero
core_matrix <- core_matrix[rowSums(core_matrix) > 0, colSums(core_matrix) > 0]

# Replace OTU names with genus names in the tax_table
tax_table_genus <- as.data.frame(tax_table(core_ps_genus))
genus_names <- tax_table_genus[rownames(core_matrix), "Genus"]

# Create a new matrix with genus names as rownames
core_matrix_genus <- core_matrix
genus_names <- gsub("g__", "", genus_names)

rownames(core_matrix_genus) <- genus_names


# Sort the matrix by row sums
row_sums <- rowSums(core_matrix_genus)
core_matrix_genus <- core_matrix_genus[order(row_sums, decreasing = TRUE), ]


# Convert the data frame to row-wise format while keeping row names
core_df_genus <- core_matrix_genus %>% as.data.frame() %>% rownames_to_column(var="Genera") %>%
  rowwise() %>%
  mutate(row_total = sum(c_across(starts_with("0.")))) %>%
  arrange(dplyr::desc(row_total)) %>%
  ungroup() %>%
  mutate(Genera = factor(Genera, levels = unique(Genera))) %>%
  dplyr::select(-row_total) %>% 
  pivot_longer(cols = -Genera,
               names_to = "prevalence",
               values_to = "value") 

# Pivoting the data longer and plotting
plot <- ggplot(core_df_genus, aes(x = prevalence, y = Genera, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "dodgerblue2", name = "Prevalence") +
  labs(title = "Core Genera at Different Prevalence Levels",
       x = "Prevalence Level",
       y = "Genera") +
  theme_minimal() +
    scale_y_discrete(limits = rev) +  # <-- flip y-axis order
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8, face = "italic"),  # <-- italic here
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )

plot


# Plot with reversed y-axis (so highest prevalence at top)
plot <- ggplot(core_df_genus, aes(x = prevalence, y = Genera, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "dodgerblue2", name = "Prevalence") +
  labs(title = "Core Genera at Different Prevalence Levels",
       x = "Prevalence Level",
       y = "Genera") +
  scale_y_discrete(limits = rev(levels(core_df_genus$Genera))) +  # flip order
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8, face = "italic"),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )

```

# 9.0 Topic modeling

## Iterate through scaling factors and topic numbers to choose the ideal combination
```{r}
iterate_res <- iterate_scaling_factors(ps_fs, 
                        "Genus",
                        scale_factors = c(10, 100, 1000, 10000), 
                        k_values= c(2, 3, 4, 5, 6), 
                        auto_chose_k = FALSE, 
                        type_column = "Type", 
                        method = "Gibbs")

iterate_res$results %>%
  mutate(
    sil_sum = sil_median_Abscess + sil_median_Plaque,
    sil_rank = min_rank(dplyr::desc(sil_sum))) %>% # Rank from highest to lowest
    arrange(sil_rank) %>% kable() 

# %>%
#     kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
#                 full_width = F,
#                 position = "center")
```

## Prep and scale input
```{r}
results_scaling <- prep_data_scale(ps_fs, "Genus", 1000, "Type")
meta_data <- results_scaling$meta_data
counts_data <- results_scaling$counts_data
```

## Use FindTopicsNumber() 
Note: There is no alpha value here
```{r}
topics <- seq(from = 2, to = 10, by = 1)
res <- RunFindTopicsNumber(counts_data, topics, "Gibbs")
res$metrics_df
```


## LDA: Create the model
You can use the info from the optimum topic numbers printed out in the above function, or adjust as needed.
```{r}
result <- create_topic_model(counts_data, 3, TRUE, "Gibbs")
model <- result$lda_model
```

## Plot beta
```{r, fig.height=3, fig.width=9}

my_colors <- c("1" = "#083D77", "2" = "#FF495C", "3" = "grey")
p <- plot_beta(result, 10, result$b_df, fill_colors = my_colors)
p
```

## Heatmap of gamma scores 
```{r, fig.height=1.5, fig.width=9}
heatmap_gamma(result, "Type", result$g_df)
```

## UMAP of gamma scores
```{r fig.height=6, fig.width=6}
UMAP_res <- plot_gamma_umap(results, "Type", colors_all, result$g_df)
```
```{r, fig.width=2.5, fig.height=3}
UMAP_res$sil_plot
```

##Membership of topic by Type
```{r, fig.height=2.5, fig.width=4}
source("MK_Microbiome_Functions.R")

topic_membership(result, "Type", colors_all, result$g_df)
```


## Heatmap of rel abundance in original data of top taxa
```{r, fig.height=3, fig.width=7}
relab_heatmap(result, ps_fs, "Genus", "Type", topic_no=1, n_top_words=10, result$b_df)

relab_heatmap(result, ps_fs, "Genus", "Type", topic_no=2, n_top_words=10, result$b_df)

relab_heatmap(result, ps_fs, "Genus", "Type", topic_no=3, n_top_words=10, result$b_df)

```



```{r}
#This is kind of a weird heatmap, I'm not sure it's actualy usful. 

posterior_results <- posterior(model)
topics_terms <- posterior_results$terms 

# Blank out most term names
term_names <- colnames(topics_terms)
term_labels <- ifelse((seq_along(term_names) %% 5) == 0, term_names, "")

# Apply these as column names
colnames(topics_terms) <- term_labels

pheatmap(log(topics_terms), 
         cluster_rows = TRUE, 
         cluster_cols = TRUE,
         scale = "none", 
         show_rownames = TRUE, 
         show_colnames = TRUE)

```



# Session Info
```{r}
sessionInfo()
```



