---
title: "02 Cohort Analysis"
author: "Madeline Krieger"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: "cosmo"
    code_folding: "show"
params:
    output_suffix: NULL

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE,
                      error = FALSE,
                      fig.keep = FALSE,
                      cache = FALSE,
                      cache.lazy = FALSE)
```


# 1.0 Load and Define

## 1.1 Load packages and presets
### 1.1.1 Packages
```{r 1.1_load-packages}
library("pacman")
pacman::p_load(
  rmarkdown, tidyverse, phyloseq, knitr, kableExtra, here, plyr, ggpubr, microViz,
  microbiome, pheatmap, vegan, reshape2, magrittr, microshades, data.table,
  Polychrome, fantaxtic, cetcolor, cowplot, MicrobiomeStat, randomForest, caret,
  mlbench, MLmetrics, mia, patchwork, digest, ANCOMBC, Maaslin2, microbiomeMarker,
  topicmodels, ldatuning, umap, tidytext, cluster, viridis, ComplexHeatmap,
  circlize, ggnewscale, stringi, corrplot, ggalluvial, Hmisc, dplyr, stringr,
  forcats, purrr, rlang, tidyr, tibble, ggplot2
)

#Create directories for output
output_path = "output/"
output_path_data = "output/data/" #Already created in script 01

# Create these directories
dir.create(output_path, showWarnings = FALSE, recursive = TRUE)
``` 


### 1.1.2 My packages
```{r}
source("MK_Microbiome_Functions.R")
```



### 1.1.2 Colors and presents
```{r}
colors_all= c("Abscess" = "#FF495C", "Plaque"="#3185FC")

core_colors = c("#FAAA00", "#3399FF","#F76F8E","#3185FC","#B8D4E3", "#FF495C","#477071", '#03CEA4',  "#5F00BA", "#BDAC9E", "white", "#FFD900")

maaslin2_colors= c("CLR_LOG" = "#2c45b5", "CLR_NONE"="#86B8FD", 
                   "CSS_LOG" = "#FF495C", "CSS_NONE"="#F2929A",
                   "TMM_LOG" = "#F4B701", "TMM_NONE"="#FEE59A",
                   "TSS_LOG" = "#8FD694", "TSS_NONE"="#D2EED4",
                   "TSS_LOGIT" = "#354F52", "TSS_AST"="#CEDDDF")

pal20 <- createPalette(22, c("#F76F8E","#03CEA4", "#3185FC"), range = c(30, 80))
pal20 <- unname(as.list(pal20))

resolution <- c("Phylum", "Class", "Order", "Family", "Genus", "Species")

set.seed(1234)
date = Sys.Date()
```


## 1.2 Load data and process

### 1.2.1 Load Data from 01
```{r}
# Load the saved Rds file
loaded_objects <- readRDS(paste0(output_path_data, "00data.rds"))

# Assign back the objects to the environment, if needed
list2env(loaded_objects, .GlobalEnv)
```


### 1.2.2 Process Phyloseq
```{r}
process_phyloseq(ps_fs)
```


# Ordinations

# Ordination parameter sweep
```{r}
result_df <- compare_ordination_methods(
  phyloseq_obj = ps_f,
  rank_transformations = c("Phylum", "Genus"),          
  trans_types = c("identity", "clr", "log", "relab"),           
  dist_cal_types = c("euclidean", "bray", "jaccard"), 
  ord_calc_methods = c("NMDS", "PCA", "PCoA"),
  No_axes_to_correlated_with_LibSize = 2,
  variance_threshold = 0.80  # Set variance threshold 
)

result_df <- result_df$results_df
```


## Organize results
```{r}
#Identify spearman_rho.* columns
rho_cols <- grep("^spearman_rho", colnames(result_df), value = TRUE)

#Compute absolute sum of all spearman_rho columns row-wise
df_with_rho_sum <- result_df %>%
  dplyr::mutate(across(all_of(rho_cols), as.numeric)) %>%
  dplyr::mutate(rho_abs_sum = apply(dplyr::select(., all_of(rho_cols)), 1, function(x) sum(abs(x), na.rm = TRUE)))

#Group, split, sort each group, and recombine
grouped_list <- df_with_rho_sum %>%
  group_by(rank_transformation, ord_calc_method) %>%
  group_split()

#Apply base R function to sort each group
sorted_list <- lapply(grouped_list, function(g) {
  g[order(-g$rho_abs_sum), ]
})

#Combine sorted groups back into one data frame
df_sorted <- bind_rows(sorted_list)

df_sorted <- df_sorted %>%
  group_by(rank_transformation, ord_calc_method) %>%
  mutate(rho_rank = row_number(rho_abs_sum)) %>%
  ungroup()

df_sorted %>% filter(rho_rank == 1 | rho_rank == 2 ) %>% kable()
```




Community separation by sample type was assessed with silhouette scores and PERMANOVA (vegan::adonis2 on Bray-Curtis distances with permutations restricted within patient pairs), with betadisper/permutest used to test whether group differences reflected unequal multivariate dispersion rather than a true compositional shift.


## Fig S1A: PCoA Species relab bray
```{r, fig.width=5, fig.height=3.5}
Species_PCoA_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Species",
  trans_type = "relab",
  ord_calc_method = "PCoA",
  dist_cal_type = "bray", 
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Species_PCoA_res$plot
Species_PCoA_res$lib_size_with_Axis1_corr

ggsave("output/Species_PCoA_relab_bray_plot.png", Species_PCoA_res$plot, width = 5, height = 3.5, dpi=600)

```

```{r}
Species_PCoA_permanova <- run_permanova_with_betadisper(
  phyloseq_obj = ps_f,
  rank_transformation = "Species",
  trans_type = "relab",
  dist_cal_type = "bray"
)
Species_PCoA_permanova$permanova
Species_PCoA_permanova$betadisper_test

```

```{r, fig.width=3, fig.height=3}
Species_PCoA_res_scores_df <- as.data.frame(Species_PCoA_res$scores)

Species_PCoA_res_scores_df <- analyze_type_clustering_on_pca(Species_PCoA_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Species_PCoA_res_scores_df$sil_plot
ggsave("output/Species_PCoA_relab_bray_sil_scores.png",Species_PCoA_res_scores_df$sil_plot, width = 3, height = 3, dpi=600)

```



## Fig 2B: PCoA Genus relab bray
```{r, fig.width=5, fig.height=3.5}
Genus_PCoA_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Genus",
  trans_type = "relab",
  ord_calc_method = "PCoA",
  dist_cal_type = "bray", 
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Genus_PCoA_res$plot
Genus_PCoA_res$lib_size_with_Axis1_corr

ggsave("output/Genus_PCoA_relab_bray_plot.png", Genus_PCoA_res$plot, width = 5, height = 3.5, dpi=600)

```

```{r}
Genus_PCoA_permanova <- run_permanova_with_betadisper(
  phyloseq_obj = ps_f,
  rank_transformation = "Genus",
  trans_type = "relab",
  dist_cal_type = "bray"
)
Genus_PCoA_permanova$permanova
Genus_PCoA_permanova$betadisper_test

```

```{r, fig.width=3, fig.height=3}
Genus_PCoA_res_scores_df <- as.data.frame(Genus_PCoA_res$scores)

Genus_PCoA_res_scores_df_sil <- analyze_type_clustering_on_pca(Genus_PCoA_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Genus_PCoA_res_scores_df_sil$sil_plot
ggsave("output/Genus_PCoA_relab_bray_sil_scores.png",Genus_PCoA_res_scores_df_sil$sil_plot, width = 3, height = 3, dpi=600)

```


## Fig 2A: PCoA Phylum relab bray
```{r, fig.width=5, fig.height=3.5}
Phylum_PCoA_res <- run_ordination_with_validation(
  phyloseq_obj = ps_f,
  rank_transformation = "Phylum",
  trans_type = "relab",
  ord_calc_method = "PCoA",
  dist_cal_type = "bray", 
  component_num = 2,
  variable = "Type",
  Axis_1 = "Axis1",
  Axis_2 = "Axis2",
  colors_list = colors_all
)

Phylum_PCoA_res$plot
Phylum_PCoA_res$lib_size_with_Axis1_corr

ggsave("output/Phylum_PCoA_relab_bray.png", Phylum_PCoA_res$plot, width = 5, height = 3.5, dpi=600)

```

```{r}
Phylum_PCoA_permanova <- run_permanova_with_betadisper(
  phyloseq_obj = ps_f,
  rank_transformation = "Phylum",
  trans_type = "relab",
  dist_cal_type = "bray"
)
Phylum_PCoA_permanova$permanova
Phylum_PCoA_permanova$betadisper_test

```

```{r, fig.width=3, fig.height=3}
Phylum_PCoA_res_scores_df <- as.data.frame(Phylum_PCoA_res$scores)

Phylum_PCoA_res_scores_df_sil <- analyze_type_clustering_on_pca(Phylum_PCoA_res_scores_df,                                           
                                          component_num = 2,
                                          colors_list = colors_all)
Phylum_PCoA_res_scores_df_sil$sil_plot

ggsave("output/Phylum_PCoA_relab_bray_sil_scores.png", Phylum_PCoA_res_scores_df_sil$sil_plot, width = 3, height = 3, dpi=600)

```


# 3.0 Fig 2C: Alpha diversity
```{r, fig.width=1.6, fig.height=3}
a_my_comparisons <- list( c("Abscess", "Plaque"))
symnum.args = list(cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), symbols = c("****", "***", "**", "*", "ns"))

Shan_df <- plot_richness(ps_f, x="Type", measures="Shannon")$data

p_Shannon <- ggplot(Shan_df, aes(x=Type, y=value, color=Type)) +
  geom_boxplot(alpha = 1, width = 0.6, outlier.shape = NA) +
  geom_jitter(size = 1.2, width = 0.15, alpha = 0.7) +
  stat_compare_means(method="wilcox.test", comparisons=a_my_comparisons, label="p.format",
                     label.y = c(5.5), 
                     vjust = -.4) +
  scale_color_manual(values = colors_all) +
  ylim(1, 6.2) +
  theme_bw(base_size = 14) +
  labs(title="Shannon") + 
  theme(
    legend.position="none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_blank(),
    strip.background = element_rect(fill="white", color="white"),
    strip.text = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )


ob_df <- plot_richness(ps_f, x="Type", measures="Observed")$data

p_Observed <- ggplot(ob_df, aes(x=Type, y=value, color=Type)) +
  geom_boxplot(alpha = 1, width = 0.6, outlier.shape = NA) +
  geom_jitter(size = 1.2, width = 0.15, alpha = 0.7) +
  stat_compare_means(method="wilcox.test", comparisons=a_my_comparisons, label="p.format", 
    label.y = c(380), 
    vjust = -.4) +
  scale_color_manual(values = colors_all) +
  ylim(0, 450) +
  theme_bw(base_size = 14) +
  labs(title="Observed") + 
  theme(
    legend.position="none",
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_blank(),
    strip.background = element_rect(fill="white", color="white"),
    strip.text = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```


```{r, fig.width=3.3, fig.height=3}
combined_alpha <- p_Shannon + p_Observed
combined_alpha

ggsave("output/Alpha_Diversity_Plots.png", combined_alpha, width = 3.3, height = 3, dpi=600)
```
```{r}

```


# 4.0 Taxa Plots

## 4.1 Average Bar Plots

### 4.1.1 Fig 3A: Phylum 
```{r taxa_phylum, fig.height=4, fig.width=4.2}

top_df <- ps_fs_t_n %>%
  tax_glom(., "Phylum") %>%
  get_top_taxa(., 10, relative = TRUE, discard_other = TRUE) %>% psmelt(.)

top_df <- top_df %>% 
  mutate(Phylum = str_remove_all(Phylum, "p__")) %>%
  mutate(Phylum = str_replace_all(Phylum, "_", " ")) %>%
  mutate(Phylum = str_remove_all(Phylum, "\\[|\\]")) %>%
  dplyr::group_by(Type,Phylum)%>%
  dplyr::summarise(Average_Abundance = mean(Abundance)) 

top_df %>%
  ggbarplot(x= "Type", y="Average_Abundance",
            fill = "Phylum",
            panel.labs.font = list(size=12),
            panel.labs.background = list(color = NULL, fill = "white"),
            font.tickslab = 14,
            font.legend = c(12),
            palette = core_colors, 
            strip.position = "top",
            color = "black",
            x.text.angle = 45, 
            y.text.angle = 0, 
            font.x = 14,
            font.y = 12,
            font.title = 16,
            rotate = FALSE,
            title = "Phlyum Level", 
            xlab = "",
            ylab = "Average Relative Abundance",
            ggtheme = theme_bw()) + 
  font("title", face="bold") +
  theme(plot.title=element_text(hjust=0.5), 
        plot.subtitle=element_text(hjust=0.5), 
        legend.text = element_text(face = "italic"), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

ggsave("output/Phylum_Average_Barplot.png", width = 4.2, height = 4, dpi=600)

```

### 4.1.2 Fig 3B: Genus 
```{r taxa_genus, fig.height=4, fig.width=3.5}

top_df <- ps_fs_t_n %>%
  tax_glom(., "Genus") %>%
  get_top_taxa(., 10, relative = TRUE, discard_other = TRUE) %>% psmelt(.)
top_df$Genus <- gsub("g__","",as.character(top_df$Genus))

top_df %>%
  dplyr::group_by(Type,Genus)%>%
  dplyr::summarise(Average_Abundance = mean(Abundance))%>%
  ggbarplot(x= "Type", y="Average_Abundance",
            fill = "Genus",
            panel.labs.font = list(size=12),
            panel.labs.background = list(color = NULL, fill = "white"),
            font.tickslab = 14,
            font.legend = c(12),
            palette = core_colors, 
            strip.position = "top",
            color = "black",
            x.text.angle = 45, 
            y.text.angle = 0, 
            font.x = 14,
            font.y = 12,
            font.title = 16,
            rotate = FALSE,
            title = "Top 10 Genera", 
            xlab = "",
            ylab = "Average Relative Abundance",
            ggtheme = theme_bw()) + 
  font("title", face="bold") +
  theme(plot.title=element_text(hjust=0.5), 
        plot.subtitle=element_text(hjust=0.5), 
        legend.text = element_text(face = "italic"), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

ggsave("output/Genus_Average_Barplot.png", width = 3.5, height = 4, dpi=600)
```


## 4.2 Heatmap
### 4.2.1 Fig 2B: Genus
```{r, fig.height=5, fig.width=6.5}
df <- ps_fs_genus_csv %>% select(-Type) %>% column_to_rownames(var="Sample") %>% as.matrix %>% t()
#df <- df[order(rowSums(df), decreasing = TRUE)[1:50], ] #select top 50

# Extract Type information from sample_data
type_annot <- data.frame(Type = ps_fs_genus_csv$Type, Sample = ps_fs_genus_csv$Sample) %>% column_to_rownames(var="Sample")
rownames(type_annot) <- colnames(df)  # Ensure match to colnames of otu_mat

# Manually define annotation colors
ann_colors <- list(Type = c(Plaque = "#3185FC", Abscess = "#FF495C"))

my_colors <- viridis(100, option = "inferno")

# Plot heatmap with annotation
pheatmap(log(df),
         scale = "none",
         clustering_distance_rows = "correlation",
         clustering_distance_cols = "correlation",
         clustering_method = "complete",
         fontsize_row = 8,
         fontsize_col = 8,
         show_rownames = FALSE,
         show_colnames = FALSE,
         annotation_col = type_annot,
         annotation_colors = ann_colors,
         color = my_colors,
         border_color = NA)

ggsave("output/Genus_Heatmap.png", width = 6.5, height = 5, dpi=600)
```


### 4.2.1 Fig S1B: Species
```{r, fig.height=5, fig.width=6.5}
df <- ps_fs_species_csv %>% select(-Type) %>% column_to_rownames(var="Sample") %>% as.matrix %>% t()
#df <- df[order(rowSums(df), decreasing = TRUE)[1:50], ] #select top 50

# Extract Type information from sample_data
type_annot <- data.frame(Type = ps_fs_species_csv$Type, Sample = ps_fs_species_csv$Sample) %>% column_to_rownames(var="Sample")
rownames(type_annot) <- colnames(df)  # Ensure match to colnames of otu_mat

# Manually define annotation colors
ann_colors <- list(Type = c(Plaque = "#3185FC", Abscess = "#FF495C"))

my_colors <- viridis(100, option = "inferno")

# Plot heatmap with annotation
p <- pheatmap(log(df),
         scale = "none",
         clustering_distance_rows = "correlation",
         clustering_distance_cols = "correlation",
         clustering_method = "complete",
         fontsize_row = 8,
         fontsize_col = 8,
         show_rownames = FALSE,
         show_colnames = FALSE,
         annotation_col = type_annot,
         annotation_colors = ann_colors,
         color = my_colors,
         filename = "./output/Species_Heatmap.png",
         height = 5,
         width = 6.5,
         border_color = NA)
p
```


# 5.0 Taxa trends
## Fig 3C, Table S2: Genus
```{r, fig.height=5, fig.width=7}
taxa_shifts_genus <- plot_taxa_shifts(top_number = 10,
                             input_df = ps_fs_genus_csv,
                             bubble_max_size = 7,
                             count_max = 25) 
taxa_shifts_genus$combined_plot

count_data_genus <- taxa_shifts_genus$bubble_data %>%
  pivot_wider(names_from = group2, values_from = count) 

write_csv(count_data_genus, "output/taxa_trends_genus_count_data.csv")

ggsave("output/Genus_Taxa_Shifts.png", taxa_shifts_genus$combined_plot, width = 7, height = 5, dpi=600)
```


## Fig 3D, Table S2: Species
```{r, fig.height=5, fig.width=12}
taxa_shifts_species <- plot_taxa_shifts(top_number = 10,
                             input_df = ps_fs_species_csv,
                             bubble_max_size = 7,
                             count_max = 25) 
taxa_shifts_species$combined_plot


count_data_species <- taxa_shifts_species$bubble_data %>%
  pivot_wider(names_from = group2, values_from = count) 


write_csv(count_data_species, "output/taxa_trends_species_count_data.csv")

ggsave("output/Species_Taxa_Shifts.png", taxa_shifts_species$combined_plot, width = 12, height = 5, dpi=600)
```



# 6.0 Differential Abundance
## 6.1 ANCOMBC2

### 6.1.1 Phylum
```{r, fig.height=7, fig.width=10}
ANCOM_ph <- runANCOM(ps_fs, 
         tax_level = "Phylum", 
         group = "Type", 
         name_of_saved_results = "../saved_analysis_files/ancombc_results_phy", 
         plot_type = "dot",
         Log2FC_cutoff = 0.5,
         plot_colors = colors_all)
```

### 6.1.2 Genus
```{r, fig.height=10, fig.width=10}
ANCOM_genus <- runANCOM(ps_fs, 
         tax_level = "Genus", 
         group = "Type", 
         name_of_saved_results = "../saved_analysis_files/ancombc_results_gen", 
         plot_type = "dot",
         plot_colors = colors_all )
```

### 6.1.2 Species
```{r, fig.height=10, fig.width=10}
ANCOM_sp <- runANCOM(ps_fs, 
         tax_level = "Species", 
         group = "Type", 
         name_of_saved_results = "../saved_analysis_files/ancombc_results_sp", 
         plot_type = "dot",
         plot_colors = colors_all )
```


## 6.2 Maaslin2

### 6.2.1 Iterate Parameters
```{r, eval=FALSE, fig.height=8, fig.width=10}
# resolutions <- c("Species", "Genus", "Phylum")
# iterate_maaslin2(ps_obj = ps_fs, 
#                  iterative_methods = iterative_methods,
#                  resolutions = resolutions,
#                  group = "Type",
#                  qval_threshold = .25, 
#                  percentage = TRUE, 
#                  plot_colors = colors_all)
```

### 6.2.2 Run Maaslin2

#### Phylum
```{r, fig.width=10, fig.height=30}
MaaslinPhylum <- run_Maaslin2(ps_obj = ps_fs, 
                                 taxa_level = "Phylum", 
                                 group = "Type",
                                 analysis_method = "LM", 
                                 normalization = "CSS",
                                 transform = "LOG", 
                                 plot_colors = colors_all,
                                 plot_type = "dot", 
                                 qval_threshold= .05)
MaaslinPhylum$results
```


#### Genus
```{r, fig.width=10, fig.height=65}
MaaslinGenus <- run_Maaslin2(ps_obj = ps_fs, 
                                 taxa_level = "Genus", 
                                 group = "Type",
                                 analysis_method = "LM", 
                                 normalization = "CSS",
                                 transform = "LOG", 
                                 plot_colors = colors_all,
                                 plot_type = "dot", 
                                 qval_threshold= .05)
MaaslinGenus$results
```


#### Species
```{r, fig.width=10, fig.height=65}
Maaslinsp <- run_Maaslin2(ps_obj = ps_fs, 
                                 taxa_level = "Species", 
                                 group = "Type",
                                 analysis_method = "LM", 
                                 normalization = "CSS",
                                 transform = "LOG", 
                                 plot_colors = colors_all,
                                 plot_type = "dot", 
                                 qval_threshold= .05)
Maaslinsp$results
```


## 6.3 AlDEX2

### 6.3.1 Iterate Parameters
https://www.nature.com/articles/s41467-022-28034-z

"...performed CLR transformation of each realization, and then performed Wilcoxon tests on the transformed realizations. The function then returned the expected Benjamini-Hochberg (BH) FDR-corrected p-value for each feature based on the results the different across Monte Carlo samples."

ALDEx2 uses by default the centred log-ratio (clr) transformation which is scale invariant

```{r, eval=FALSE,  fig.width=20, fig.height=12}
# resolutions <- c("Species", "Genus", "Phylum")
#   
# iterate_aldex2(ps_obj = ps_fs, 
#                  iterative_methods = iterative_methods,
#                  resolutions = resolutions,
#                  group = "Type",
#                  percentage = FALSE)
# 
# aldex2_iterative_summary_table %>%
#   filter(resolution == "Genus") %>%
#   filter(paired == FALSE) %>%
#   filter(paired == FALSE) %>%
#   filter(denom == "all") %>%
#   filter(transform == "log10") %>%
#   dplyr::arrange(desc(significant_features)) %>%
#   filter(significant_features > 0)
```

### 6.3.2 Run Aldex2

#### Phylum
```{r, fig.width=10, fig.height=3}
ALDEX_ph <- run_aldex2(ps_fs, 
                  "Type", 
                  "Phylum",
                  method = "wilcox.test",
                  transform = "log10",
                  normalization = "CSS", 
                 plot_colors = colors_all)
ALDEX_ph$results
```


#### Genus
```{r, fig.width=7, fig.height=5}
ALDEX_genus <- run_aldex2(ps_fs, 
                  "Type", 
                  "Genus",
                  method = "wilcox.test",
                  transform = "log10",
                  normalization = "CSS", 
                 plot_colors = colors_all)
ALDEX_genus$result
```


#### Species
```{r, fig.width=7, fig.height=5}
ALDEX_sp <- run_aldex2(ps_fs, 
                  "Type", 
                  "Species",
                  method = "wilcox.test",
                  transform = "log10",
                  normalization = "CSS", 
                 plot_colors = colors_all)
```


## 6.4 Combine results

### Fig 4A: Phylum
```{r, fig.height=3.6, fig.width=3.2}
DA_results_df_Phylum <- combine_DA(maaslin2_results_Phylum_0.05, ancombc2_results_df_Phylum, aldex2_res_Phylum, "Type", "Phylum")


DA_results_df_Phylum$results

write_csv(DA_results_df_Phylum$results, "output/phylum_DA_combined_results.csv")
```

```{r, fig.width=6.2, fig.height=3}
phylum_combined_bubbles_P <- combine_DA_with_bubbles(da = DA_results_df_Phylum,
  CSV_formatted_relab_df = ps_fs_phylum_csv,
  group = "Type",               # your contrast label used in ANCOMBC2 column names
  tax_level = "Phylum",
  heatmap_side = "Plaque",        # or "Abscess"
  bubble_max_size = 9,
  plot_layout_widths = c(2, 1))

phylum_combined_bubbles_P$combined_plot
ggsave("output/Phylum_DA_Bubble_Plaque_labs.png", phylum_combined_bubbles_P$combined_plot, width = 6.2, height = 3, dpi=600)

phylum_bubbledata_P <- phylum_combined_bubbles_P$bubble_data %>% 
  pivot_wider(names_from = group2, values_from = count) %>%
  dplyr::mutate(Type = "Plaque")
```

```{r, fig.width=3.2, fig.height=1.1}
phylum_combined_bubbles_P$combined_plot_nolabs

ggsave("output/Phylum_DA_Bubble_Plaque_no-labs.png", phylum_combined_bubbles_P$combined_plot_nolabs, width = 3.2, height = 1.1, dpi=600)

```


```{r, fig.width=6, fig.height=3.4}
phylum_combined_bubbles_A <- combine_DA_with_bubbles(da = DA_results_df_Phylum,
  CSV_formatted_relab_df = ps_fs_phylum_csv,
  group = "Type",               # your contrast label used in ANCOMBC2 column names
  tax_level = "Phylum",
  heatmap_side = "Abscess",        
  bubble_max_size = 9,
  plot_layout_widths = c(2, 1))
phylum_combined_bubbles_A$combined_plot

ggsave("output/Phylum_DA_Bubble_Abscess_labs.png", phylum_combined_bubbles_A$combined_plot, width = 6, height = 3.4, dpi=600)

phylum_bubbledata_A <- phylum_combined_bubbles_A$bubble_data %>% 
  pivot_wider(names_from = group2, values_from = count) %>%
  dplyr::mutate(Type = "Abscess")
```



```{r, fig.width=3.8, fig.height=1.45}
phylum_combined_bubbles_A$combined_plot_nolabs

ggsave("output/Phylum_DA_Bubble_Abscess_no-labs.png", phylum_combined_bubbles_A$combined_plot_nolabs, width = 3.8, height = 1.45, dpi=600)
```

```{r}
# Combine the count data results and write to a file
phylum_count_data <- phylum_bubbledata_A %>%
  dplyr::full_join(phylum_bubbledata_P)

write_csv(phylum_count_data, "output/phylum_count_data_DA_bubbledata.csv")
```


### Fig 4B: Genus


```{r, fig.height=12, fig.width=3.9}
DA_results_df_genus <- combine_DA(maaslin2_results_Genus_0.05, ancombc2_results_df_Genus, aldex2_res_Genus, "Type", "Genus", filter_confidence = NULL)

DA_results_df_genus$results

write_csv(DA_results_df_genus$results, "output/genus_DA_combined_results.csv")
```

```{r, fig.width=7.5, fig.height=8}
genus_combined_bubbles_P <- combine_DA_with_bubbles(da = DA_results_df_genus,
  CSV_formatted_relab_df = ps_fs_genus_csv,
  group = "Type",               # your contrast label used in ANCOMBC2 column names
  tax_level = "genus",
  heatmap_side = "Plaque",        # or "Abscess"
  bubble_max_size = 9,
  plot_layout_widths = c(2, 1))
genus_combined_bubbles_P$combined_plot

ggsave("output/Genus_DA_Bubble_Plaque_labs.png", genus_combined_bubbles_P$combined_plot, width = 7.5, height = 8, dpi=600)

genus_bubbledata_P <- genus_combined_bubbles_P$bubble_data %>% 
  pivot_wider(names_from = group2, values_from = count) %>%
  dplyr::mutate(Type = "Plaque")
```

```{r,  eval=FALSE, fig.width=11.5, fig.height=9}
genus_combined_bubbles_P$combined_plot_nolabs
```

```{r, fig.width=7.9, fig.height=6.8}

genus_combined_bubbles_A <- combine_DA_with_bubbles(da = DA_results_df_genus,
  CSV_formatted_relab_df = ps_fs_genus_csv,
  group = "Type",               
  tax_level = "genus",
  heatmap_side = "Abscess",        
  bubble_max_size = 9,
  plot_layout_widths = c(2, 1))
genus_combined_bubbles_A$combined_plot

ggsave("output/Genus_DA_Bubble_Abscess_labs.png", genus_combined_bubbles_A$combined_plot, width = 7.9, height = 6.8, dpi=600)

genus_bubbledata_A <- genus_combined_bubbles_A$bubble_data %>% 
  pivot_wider(names_from = group2, values_from = count) %>%
  dplyr::mutate(Type = "Abscess")
```

```{r,  eval=FALSE,  fig.width=12, fig.height=7}
genus_combined_bubbles_A$combined_plot_nolabs
```

```{r}
# Combine the count data results and write to a file
genus_count_data <- genus_bubbledata_A %>%
  dplyr::full_join(genus_bubbledata_P)

write_csv(genus_count_data, "output/genus_count_data_DA_bubbledata.csv")
```



### Fig 5: Species
```{r, fig.height=11, fig.width=5.5}
DA_results_df_sp <- combine_DA(maaslin2_results_Species_0.05, ancombc2_results_df_Species, aldex2_res_Species, "Type", "Species")

DA_results_df_sp$results

write_csv(DA_results_df_sp$results, "output/species_DA_combined_results.csv")
```

```{r, fig.width=10.6, fig.height=11.1}
sp_combined_bubbles_P <- combine_DA_with_bubbles(da = DA_results_df_sp,
  CSV_formatted_relab_df = ps_fs_species_csv,
  group = "Type",               
  tax_level = "species",
  heatmap_side = "Plaque",        
  bubble_max_size = 7,
  plot_layout_widths = c(2, 1))
sp_combined_bubbles_P$combined_plot

ggsave("output/Species_DA_Bubble_Plaque_labs.png", sp_combined_bubbles_P$combined_plot, width = 10.6, height = 11.1, dpi=600)

sp_bubbledata_P <- sp_combined_bubbles_P$bubble_data %>% 
  pivot_wider(names_from = group2, values_from = count) %>%
  dplyr::mutate(Type = "Plaque")


```


```{r, fig.width=3.5, fig.height=12}
sp_combined_bubbles_P$combined_plot_nolabs
```

```{r, fig.width=10.9, fig.height=7.2}
species_combined_bubbles_A <- combine_DA_with_bubbles(da = DA_results_df_sp,
  CSV_formatted_relab_df = ps_fs_species_csv,
  group = "Type",               
  tax_level = "species",
  heatmap_side = "Abscess",        
  bubble_max_size = 9,
  plot_layout_widths = c(2, 1))
species_combined_bubbles_A$combined_plot

ggsave("output/Species_DA_Bubble_Abscess_labs.png", species_combined_bubbles_A$combined_plot, width = 10.9, height = 7.2, dpi=600)

sp_bubbledata_A <- species_combined_bubbles_A$bubble_data %>% 
  pivot_wider(names_from = group2, values_from = count) %>%
  dplyr::mutate(Type = "Abscess")


```


```{r, fig.width=3.5, fig.height=6}
species_combined_bubbles_A$combined_plot_nolabs
```

```{r}
# Combine the count data results and write to a file
sp_count_data <- sp_bubbledata_A %>%
  dplyr::full_join(sp_bubbledata_P)

write_csv(sp_count_data, "output/species_count_data_DA_bubbledata.csv")

```

## 6.5 Per taxa plots

### Fig S3A: Fusobacteria phylum
```{r, fig.width=2.5, fig.height=2.8}
plot_all_taxa(ps_obj = ps_fs,
              transformation = "TSS",
              group = "Type",
              taxa_level = "Phylum",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("p__Fusobacteria" ))

ggsave("output/Selected_Phyla_Fusobacteria_Dotplot_TSS.png", width = 2.5, height = 2.8, dpi=600)
```
```{r, fig.width=2.5, fig.height=2.8}
plot_all_taxa(ps_obj = ps_fs,
              transformation = "CLR",
              group = "Type",
              taxa_level = "Phylum",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("p__Fusobacteria" ))

ggsave("output/Selected_Phyla_Fusobacteria_Dotplot_CLR.png", width = 2.5, height = 2.8, dpi=600)
```

### Fig S2A-B: Selected Genus Plots
```{r, fig.width=8, fig.height=3}
plot_all_taxa(ps_obj = ps_fs,
              transformation = "TSS",
              group = "Type",
              taxa_level = "Genus",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c( "g__Actinomyces", "g__Corynebacterium", "g__Leptotrichia","g__Dialister", "g__Treponema",  "g__Veillonellaceae_[G-1]"))

ggsave("output/Selected_Genus_Dotplot_1.png", width = 8, height = 3, dpi=600)
```

```{r, fig.width=8, fig.height=3}
plot_all_taxa(ps_obj = ps_fs,
              transformation = "TSS",
              group = "Type",
              taxa_level = "Genus",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("g__Lautropia", "g__Neisseria",  "g__Rothia", "g__Lancefieldella", "g__Oribacterium", "g__Solobacterium"))

ggsave("output/Selected_Genus_Dotplot_2.png", width = 8, height = 3, dpi=600)
```

### Fig S3B: Fusobacterium genus
```{r, fig.width=2.5, fig.height=3}
clr_fuso <- plot_all_taxa(ps_obj = ps_fs,
              transformation = "CLR",
              group = "Type",
              taxa_level = "Genus",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("g__Fusobacterium"))
clr_fuso
ggsave("output/Selected_Genus_CLR_Fusobacterium_Dotplot.png", clr_fuso, width = 2.5, height = 3, dpi=600)

tss_fuso <- plot_all_taxa(ps_obj = ps_fs,
              transformation = "TSS",
              group = "Type",
              taxa_level = "Genus",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("g__Fusobacterium"))
tss_fuso
ggsave("output/Selected_Genus_TSS_Fusobacterium_Dotplot.png", tss_fuso, width = 2.5, height = 3, dpi=600)

```

### Fig S3C-D: Selected Fusobacterium Species Plots
```{r, fig.width=4, fig.height=3}
sp_fuso_tss <- plot_all_taxa(ps_obj = ps_fs,
              transformation = "TSS",
              group = "Type",
              taxa_level = "Species",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("g__Fusobacterium_s__nucleatum", "g__Fusobacterium_s__periodonticum"))
sp_fuso_tss
ggsave("output/Selected_Species_TSS_Fusobacterium_Dotplot.png", sp_fuso_tss, width = 4, height = 3, dpi=600)

sp_fuso_clr <- plot_all_taxa(ps_obj = ps_fs,
              transformation = "CLR",
              group = "Type",
              taxa_level = "Species",
              plot_type = "dot",
              plot_colors = colors_all,
              taxa_filter = c("g__Fusobacterium_s__nucleatum", "g__Fusobacterium_s__periodonticum"))
sp_fuso_clr
ggsave("output/Selected_Species_CLR_Fusobacterium_Dotplot.png", sp_fuso_clr, width = 4, height = 3, dpi=600)
```



# 7.0 Random Forest: Genus
```{r}
set.seed(1234)

df <- ps_fs_genus_csv %>% column_to_rownames(var="Sample")
k <- 4

# Create stratified folds (list of test indices for each fold)
folds <- createFolds(df$Type, k = k, returnTrain = TRUE)

# View class distribution in each fold
for (i in seq_along(folds)) {
  cat("Fold", i)
  print(table(df[folds[[i]], "Type"]))
  cat("\n")
}

# Set up trainControl with custom indices
fitControl <- trainControl(
  method = "cv",
  number = k,
  index = folds,  # custom folds for training
  classProbs = TRUE,
  savePredictions = TRUE,
  summaryFunction = twoClassSummary
)

# Train the random forest model
set.seed(123)
model <- caret::train(
  Type ~ .,
  data = df,
  method = "rf",
  trControl = fitControl,
  metric = "ROC"
)


```


### Fig 6C: Rank features by fold and create heatmap
```{r, fig.width=3.8, fig.height=4}

if ("package:plyr" %in% search()) {
  detach("package:plyr", unload = TRUE, character.only = TRUE)
}

set.seed(123) 
rank_table <- rank_rf_feature_table(df, folds, top_n = 10, add_mean = TRUE)%>%
    mutate(Feature = str_remove_all(Feature, "g__")) %>%
  mutate(Feature = str_replace_all(Feature, "_", " ")) %>%
  mutate(Feature = str_remove_all(Feature, "\\[|\\]")) 


# Example: show the first few rows
rank_table %>% kable()


mat <- rank_table %>%
  select(-Feature, -MeanRank) %>%
  as.data.frame() 


rownames(mat) <- rank_table$Feature
rownames(mat) <- sub("`","", rownames(mat)) #Need this here twice
rownames(mat) <- sub("`","", rownames(mat)) 

rownames(mat) <- sub("^g__","", rownames(mat))

labs <- rownames(mat)

# Make taxa labels italics
labs_expr <- parse(text = sprintf("italic('%s')",
                                  gsub("'", "\\\\'", labs, fixed = TRUE)))

p <- pheatmap(log(mat),
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         labels_row = labs_expr,
         color = colorRampPalette(c("#752ACB", "white", "#C0E6DE"))(50),
        display_numbers = round(as.matrix(mat), 1),  # 1 decimal place
        fontsize_number = 10,
         main = "Feature Ranks per Fold")
p 

#ggsave("output/RF_Feature_Ranks_per_Fold_Heatmap.png", p, width = 3.8, height = 4, dpi=600)
```


### Fig 6B: ROC Curve
```{r, fig.width=4, fig.height=3}
#Plot ROC Curve for each fold of the model
for_lift <- data.frame(Class = model$pred$obs, rf = model$pred$Plaque, resample = model$pred$Resample)
lift_df <-  data.frame()

for (fold in unique(for_lift$resample)) {
  fold_df <- dplyr::filter(for_lift, resample == fold)
  lift_obj_data <- lift(Class ~ rf, data = fold_df, class = "Plaque")$data
  lift_obj_data$fold = fold
  lift_df = rbind(lift_df, lift_obj_data)
}

lift_obj <- lift(Class ~ rf, data = for_lift, class = "Plaque")

roc_p <- ggplot(lift_df) +
  geom_line(aes(x = 1 - Sp, y = Sn, color = fold)) +
  scale_color_manual(
    name = "Fold",
    values = c("#DD1155", "#00CC99", "#F4B942", "#3399FF")
  ) +
  theme_bw()
roc_p

ggsave("output/RF_ROC_Curve_per_Fold.png", roc_p, width = 4, height = 3, dpi=600)
```



### Print AUC Values
```{r}
library("plyr")
AUC_score <-ddply(model$pred, "Resample", summarise,
      accuracy = Accuracy(pred, obs))

AUC_score %>% kbl() %>%
  kable_minimal()


if ("package:plyr" %in% search()) {
  detach("package:plyr", unload = TRUE, character.only = TRUE)
}
```


### Print out top features
```{r}
# Extract variable importance
importance <- varImp(model, scale = FALSE)

# Convert to data frame for easier manipulation
importance_df <- as.data.frame(importance$importance)
importance_df$Feature <- rownames(importance_df)

# Print top 10 most important features
top_features <- importance_df %>%
  arrange(dplyr::desc(Overall)) %>%
  filter(Overall > 0 )

kable(top_features, format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                position = "center")
```




# 7.0 Random Forest: Species
```{r}
set.seed(1234)

df <- ps_fs_species_csv %>% column_to_rownames(var="Sample")
k <- 4

# Create stratified folds (list of test indices for each fold)
folds <- createFolds(df$Type, k = k, returnTrain = TRUE)

# View class distribution in each fold
for (i in seq_along(folds)) {
  cat("Fold", i)
  print(table(df[folds[[i]], "Type"]))
  cat("\n")
}

# Set up trainControl with custom indices
fitControl <- trainControl(
  method = "cv",
  number = k,
  index = folds,  # custom folds for training
  classProbs = TRUE,
  savePredictions = TRUE,
  summaryFunction = twoClassSummary
)

# Train the random forest model
set.seed(123)
model <- caret::train(
  Type ~ .,
  data = df,
  method = "rf",
  trControl = fitControl,
  metric = "ROC"
)


```


### Fig S4B: Rank features by fold and create heatmap
```{r, fig.width=5, fig.height=5}

if ("package:plyr" %in% search()) {
  detach("package:plyr", unload = TRUE, character.only = TRUE)
}

set.seed(123) 
rank_table <- rank_rf_feature_table(df, folds, top_n = 10, add_mean = TRUE)%>%
  mutate(
    Feature = str_remove_all(Feature, "g__"),
    Feature = str_replace_all(Feature, "_", " "),
    Feature = str_remove_all(Feature, "\\[|\\]"),
    Feature  = str_replace_all(Feature, " s  ", " "))


# Example: show the first few rows
rank_table %>% kable()


mat <- rank_table %>%
  select(-Feature, -MeanRank) %>%
  as.data.frame() 


rownames(mat) <- rank_table$Feature
rownames(mat) <- sub("`","", rownames(mat)) #Need this here twice
rownames(mat) <- sub("`","", rownames(mat)) 

rownames(mat) <- sub("^g__","", rownames(mat))

labs <- rownames(mat)

# Make taxa labels italics
labs_expr <- parse(text = sprintf("italic('%s')",
                                  gsub("'", "\\\\'", labs, fixed = TRUE)))

p <- pheatmap(
  log(mat),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  labels_row = labs_expr,
  color = colorRampPalette(c("#752ACB", "white", "#C0E6DE"))(50),
  display_numbers = round(as.matrix(mat), 1),
  fontsize_number = 10,
  main = "Feature Ranks per Fold",
  filename = "output/RF_Feature_Ranks_per_Fold_Heatmap_species.png",
  width = 5,
  height = 5
)

p
```


### Fig S4A: ROC Curve
```{r, fig.width=4, fig.height=3}
#Plot ROC Curve for each fold of the model
for_lift <- data.frame(Class = model$pred$obs, rf = model$pred$Plaque, resample = model$pred$Resample)
lift_df <-  data.frame()

for (fold in unique(for_lift$resample)) {
  fold_df <- dplyr::filter(for_lift, resample == fold)
  lift_obj_data <- lift(Class ~ rf, data = fold_df, class = "Plaque")$data
  lift_obj_data$fold = fold
  lift_df = rbind(lift_df, lift_obj_data)
}

lift_obj <- lift(Class ~ rf, data = for_lift, class = "Plaque")

roc_p <- ggplot(lift_df) +
  geom_line(aes(x = 1 - Sp, y = Sn, color = fold)) +
  scale_color_manual(
    name = "Fold",
    values = c("#DD1155", "#00CC99", "#F4B942", "#3399FF")
  ) +
  theme_bw()
roc_p

ggsave("output/RF_ROC_Curve_per_Fold_species.png", roc_p, width = 4, height = 3, dpi=600)
```



### Print AUC Values
```{r}
library("plyr")
AUC_score <-ddply(model$pred, "Resample", summarise,
      accuracy = Accuracy(pred, obs))

AUC_score %>% kbl() %>%
  kable_minimal()


if ("package:plyr" %in% search()) {
  detach("package:plyr", unload = TRUE, character.only = TRUE)
}
```


### Print out top features
```{r}
# Extract variable importance
importance <- varImp(model, scale = FALSE)

# Convert to data frame for easier manipulation
importance_df <- as.data.frame(importance$importance)
importance_df$Feature <- rownames(importance_df)

# Print top 10 most important features
top_features <- importance_df %>%
  arrange(dplyr::desc(Overall)) %>%
  filter(Overall > 0 )

kable(top_features, format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                position = "center")
```



# 8.0 Core microbiome
## Core Microbiome
```{r}
core_ps <- ps_fs
core_ps_n <- transform_sample_counts(core_ps, function(x) (x) / sum(x))

# Agglomerate to the genus level
core_ps_genus <- tax_glom(core_ps, taxrank = "Genus")

# Initialize an empty dataframe to store results
results <- data.frame(prevalence = numeric(), percent_OTUs = numeric())

# Get the total number of OTUs at the species level
total_OTUs <- length(taxa_names(core_ps_genus))

# Loop through prevalence values from 0.1 to 1
for (p in seq(0.1, 1, by = 0.1)) {
  core.taxa <- core_members(core_ps_genus, detection = 0.00001, prevalence = p)
  num_OTUs <- length(core.taxa)
  percent_OTUs <- (num_OTUs / total_OTUs) * 100
  
  # Append the results to the dataframe
  results <- rbind(results, data.frame(prevalence = p*100, percent_OTUs = percent_OTUs))
}

# Create a line graph using ggplot2
ggplot(results, aes(x = prevalence, y = percent_OTUs)) +
  geom_line() +
  geom_point() +
  labs(title = "Percent of Total Taxa Detectioned",
       x = "Percent of Samples Containing Taxa",
       y = "Percent of Total Taxa") +
  theme_minimal()

```

## Core microbiome heatmap
```{r, fig.height=5, fig.width=5}

# Get the total list of species
genus_list <- taxa_names(core_ps_genus) 

# Initialize an empty list to store core genera at each prevalence level
core_genus_list <- list()

# Loop through prevalence values 
for (p in seq(0.5, 1, by = 0.05)) {
  core.taxa <- core_members(core_ps_genus, detection = 1, prevalence = p)
  core_genus_list[[as.character(p)]] <- core.taxa
}

# Create a binary matrix indicating the presence (1) or absence (0) of genera in the core microbiome
core_matrix <- matrix(0, nrow = length(genus_list), ncol = length(core_genus_list),
                      dimnames = list(genus_list, names(core_genus_list)))

for (p in names(core_genus_list)) {
  core_matrix[core_genus_list[[p]], p] <- 1
}

# Remove rows and columns that sum to zero
core_matrix <- core_matrix[rowSums(core_matrix) > 0, colSums(core_matrix) > 0]

# Replace OTU names with genus names in the tax_table
tax_table_genus <- as.data.frame(tax_table(core_ps_genus))
genus_names <- tax_table_genus[rownames(core_matrix), "Genus"]

# Create a new matrix with genus names as rownames
core_matrix_genus <- core_matrix
genus_names <- gsub("g__", "", genus_names)

rownames(core_matrix_genus) <- genus_names


# Sort the matrix by row sums
row_sums <- rowSums(core_matrix_genus)
core_matrix_genus <- core_matrix_genus[order(row_sums, decreasing = TRUE), ]


# Convert the data frame to row-wise format while keeping row names
core_df_genus <- core_matrix_genus %>% as.data.frame() %>% rownames_to_column(var="Genera") %>%
  rowwise() %>%
  mutate(row_total = sum(c_across(starts_with("0.")))) %>%
  arrange(dplyr::desc(row_total)) %>%
  ungroup() %>%
  mutate(Genera = factor(Genera, levels = unique(Genera))) %>%
  dplyr::select(-row_total) %>% 
  pivot_longer(cols = -Genera,
               names_to = "prevalence",
               values_to = "value") 

# Pivoting the data longer and plotting
plot <- ggplot(core_df_genus, aes(x = prevalence, y = Genera, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "#564592", name = "Prevalence") +
  labs(title = "Core Genera at Different Prevalence Levels",
       x = "Prevalence Level",
       y = "Genera") +
  theme_minimal() +
    scale_y_discrete(limits = rev) +  # <-- flip y-axis order
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8, face = "italic"),  # <-- italic here
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )

plot


# Plot with reversed y-axis (so highest prevalence at top)
plot <- ggplot(core_df_genus, aes(x = prevalence, y = Genera, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "dodgerblue2", name = "Prevalence") +
  labs(title = "Core Genera at Different Prevalence Levels",
       x = "Prevalence Level",
       y = "Genera") +
  scale_y_discrete(limits = rev(levels(core_df_genus$Genera))) +  # flip order
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8, face = "italic"),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )

ggsave("output/Core_Genera_Heatmap.png", plot, width = 5, height = 5, dpi=600)

```

# 9.0 Topic modeling: Genus

## Iterate through scaling factors and topic numbers to choose the ideal combination
```{r, eval=F}
iterate_res <- iterate_scaling_factors(ps_fs, 
                        "Genus",
                        scale_factors = c(10, 100, 1000, 10000), 
                        k_values= c(2, 3, 4, 5, 6), 
                        auto_chose_k = FALSE, 
                        type_column = "Type", 
                        method = "Gibbs")

iterate_res$results %>%
  mutate(
    sil_sum = sil_median_Abscess + sil_median_Plaque,
    sil_rank = min_rank(dplyr::desc(sil_sum))) %>% # Rank from highest to lowest
    arrange(sil_rank) %>% kable() 

# %>%
#     kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
#                 full_width = F,
#                 position = "center")
```

## Prep and scale input
```{r}
results_scaling_genus <- prep_data_scale(ps_fs, "Genus", 1000, "Type")
meta_data <- results_scaling_genus$meta_data
counts_data_genus <- results_scaling_genus$counts_data
```

## Fig S5B: Use FindTopicsNumber() 
Note: There is no alpha value here
```{r, fig.widht=4, fig.height=4}
topics <- seq(from = 2, to = 10, by = 1)
res_genus <- RunFindTopicsNumber(counts_data_genus, topics, "Gibbs")

metric_plot_genus <- res_genus$plot + 
  geom_vline(xintercept = 7, linetype = "dashed", linewidth = 0.8, color = "red") +
  theme_bw(base_size=13) +
  theme(
  axis.text.x = element_text(size=13, color="black"),
  strip.background = element_rect(fill = "white", color = "black"),
  strip.text = element_text(size = 14, face = "bold"), 
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank()) + 
 labs(title = "Genus Level Metrics") +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5)
  )



ggsave("output/Topic_Model_FindTopicsNumber_Genus.png", metric_plot_genus, width = 6, height = 4, dpi=600)

res_genus$metrics_df
```


## LDA: Create the model
You can use the info from the optimum topic numbers printed out in the above function, or adjust as needed.

```{r}
result_genus <- create_topic_model(counts_data_genus, 7, TRUE, "Gibbs")
model_genus <- result_genus$lda_model_genus
```


## Fig S8A: Membership of topic by Type
```{r, fig.height=2.5, fig.width=8.4}

topic_membership_res_genus <- topic_membership(result_genus, "Type", colors_all, result_genus$g_df)

hist(result_genus$g_df$gamma)

plot_genus <- topic_membership_res_genus$plot + 
  stat_compare_means(
    comparisons = list(c("Abscess", "Plaque")),
    method = "wilcox.test",
    label.y = c(1), 
    vjust = .45,
    label = "p.signif")+ 
    ylim(0, 1.1)

plot_genus <- plot_genus + scale_x_discrete(labels = c("Abscess" = "A", "Plaque" = "P")) + 
  theme (
    axis.text=element_text(size=13, color="black"),
  )

plot_genus

ggsave("output/Topic_Model_Topic_Membership_by_Type_Genus.png", plot_genus, width = 8.4, height = 2.5, dpi=600)
```


## Fig S8C: Plot beta
```{r, fig.height=3, fig.width=20}
### Plot beta
#plot_beta(result, 15)
#Plot beta, or the numbers that are assigned to each word in a topic. If the beta score is higher, the word matters more in that topic.
result_genus$b_df$topic <- factor(result_genus$b_df$topic, levels = c("1", "4", "6", "3", "5", "2", "7"))


my_colors <- c("1" = "#3185FC", "4" = "#3185FC","6" = "#3185FC","3" = "#FF495C", "5" = "#FF495C")
p_genus <- plot_beta(result_genus, 10, result_genus$b_df,  fill_colors = my_colors)
p_genus

ggsave("output/Topic_Model_Beta_Barplots_Genus.png", p_genus, width = 20, height = 3, dpi=600)
```



## Fig S8B: Heatmap of gamma scores 
```{r, fig.height=2.7, fig.width=9}
heatmap_gamma_genus <- heatmap_gamma(result_genus, "Type", result_genus$g_df)
png("output/Topic_Model_Gamma_Heatmap_Genus.png",
    width = 9, height = 2, units = "in", res = 600)

ComplexHeatmap::draw(heatmap_gamma_genus)  # <- explicitly use this one

dev.off()
```

## UMAP of gamma scores
```{r fig.height=6, fig.width=6}
UMAP_res_genus <- plot_gamma_umap(results_genus, "Type", colors_all, result_genus$g_df)
UMAP_res_genus$umap

ggsave("output/Topic_Model_Gamma_UMAP_Genus.png", UMAP_res_genus$umap, width = 5, height = 4, dpi=600)
```

```{r, fig.width=3, fig.height=3}
UMAP_res_genus$sil_plot

ggsave("output/Topic_Model_Gamma_UMAP_Silhouette_Genus.png", UMAP_res_genus$sil_plot, width = 3, height = 3, dpi=600)
```


## Fig S9: Relab heatmaps by topic
```{r, fig.height=2, fig.width=7}
topic1_res_genus <- heatmap_by_topic(result_genus$b_df, ps_fs_genus_csv, n_top = 10, topic = 1)

png("output/Topic_Model_RelAb_Heatmap_Topic1_genus.png",
    width = 6.5, height = 2.5, units = "in", res = 600)

ComplexHeatmap::draw(topic1_res_genus$heatmap)

dev.off()

topic1_res_genus$heatmap
```

```{r, fig.height=2, fig.width=7.5}
topic2_res_genus <- heatmap_by_topic(result_genus$b_df, ps_fs_genus_csv, n_top = 10, topic = 2)

png("output/Topic_Model_RelAb_Heatmap_Topic2_genus.png",
    width = 6.5, height = 2.5, units = "in", res = 600)

ComplexHeatmap::draw(topic2_res_genus$heatmap)  # <- explicitly use this one

dev.off()

topic2_res_genus$heatmap

```

```{r, fig.height=2.5, fig.width=6.5}
topic3_res_genus <- heatmap_by_topic(result_genus$b_df, ps_fs_genus_csv, n_top = 10, topic = 3)

png("output/Topic_Model_RelAb_Heatmap_Topic3_genus.png",
    width = 6.5, height = 2.5, units = "in", res = 600)

ComplexHeatmap::draw(topic3_res_genus$heatmap)  # <- explicitly use this one

dev.off()

topic3_res_genus$heatmap
```

```{r, fig.height=2, fig.width=7}
topic4_res_genus <- heatmap_by_topic(result_genus$b_df, ps_fs_genus_csv, n_top = 10, topic = 4)

png("output/Topic_Model_RelAb_Heatmap_Topic4_genus.png",
    width = 6.5, height = 2.5, units = "in", res = 600)

ComplexHeatmap::draw(topic4_res_genus$heatmap)  # <- explicitly use this one

dev.off()

topic4_res_genus$heatmap

```

```{r, fig.height=2, fig.width=7}
topic5_res_genus <- heatmap_by_topic(result_genus$b_df, ps_fs_genus_csv, n_top = 10, topic = 5)


png("output/Topic_Model_RelAb_Heatmap_Topic5_genus.png",
    width = 6.5, height = 2.5, units = "in", res = 600)

ComplexHeatmap::draw(topic5_res_genus$heatmap)

dev.off()

topic5_res_genus$heatmap
```

```{r, fig.height=2, fig.width=7}
topic6_res_genus <- heatmap_by_topic(result_genus$b_df, ps_fs_genus_csv, n_top = 10, topic = 6)


png("output/Topic_Model_RelAb_Heatmap_Topic6_genus.png",
    width = 6.5, height = 2.5, units = "in", res = 600)

ComplexHeatmap::draw(topic6_res_genus$heatmap)

dev.off()

topic6_res_genus$heatmap
```

```{r, fig.height=2, fig.width=7}
topic7_res_genus <- heatmap_by_topic(result_genus$b_df, ps_fs_genus_csv, n_top = 10, topic = 7)


png("output/Topic_Model_RelAb_Heatmap_Topic7_genus.png",
    width = 6.5, height = 2.5, units = "in", res = 600)

ComplexHeatmap::draw(topic7_res_genus$heatmap)

dev.off()

topic7_res_genus$heatmap
```


# 9.0 Topic modeling: Species

## Iterate through scaling factors and topic numbers to choose the ideal combination
```{r, eval=F}
iterate_res <- iterate_scaling_factors(ps_fs, 
                        "Species",
                        scale_factors = c(10, 100, 1000), 
                        k_values= c(2, 3, 4, 5, 6), 
                        auto_chose_k = FALSE, 
                        type_column = "Type", 
                        method = "Gibbs")

iterate_res$results %>%
  mutate(
    sil_sum = sil_median_Abscess + sil_median_Plaque,
    sil_rank = min_rank(dplyr::desc(sil_sum))) %>% # Rank from highest to lowest
    arrange(sil_rank) %>% kable() 

# %>%
#     kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
#                 full_width = F,
#                 position = "center")
```

## Prep and scale input
```{r}
results_scaling_species <- prep_data_scale(ps_fs, "Species", 1000, "Type")
meta_data <- results_scaling_species$meta_data
counts_data_species <- results_scaling_species$counts_data
```

## Fig S5A: Use FindTopicsNumber() 
```{r}
topics <- seq(from = 2, to = 10, by = 1)
res_species <- RunFindTopicsNumber(counts_data_species, topics, "Gibbs")

metric_plot_species <- res_species$plot +
  geom_vline(xintercept = 6, linetype = "dashed", linewidth = 0.8, color = "red") +
  theme_bw(base_size=13) +
  theme(
  axis.text.x = element_text(size=13, color="black"),
  strip.background = element_rect(fill = "white", color = "black"),
  strip.text = element_text(size = 14, face = "bold"), 
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank()) + 
 labs(title = "Species Level Metrics") +
  theme(
    plot.title = element_text(size = 18, face = "bold", hjust = 0.5)
  )
metric_plot_species

ggsave("output/Topic_Model_FindTopicsNumber_Species.png", metric_plot_species, width = 6, height = 4, dpi=600)

res_species$metrics_df
```


## LDA: Create the model
You can use the info from the optimum topic numbers printed out in the above function, or adjust as needed.

```{r}
result_species <- create_topic_model(counts_data_species, 6, TRUE, "Gibbs")
model_species <- result_species$lda_model
```

## Fig 7C: Membership of topic by Type
```{r, fig.height=2.5, fig.width=7.4}

topic_membership_res_species <- topic_membership(result_species, "Type", colors_all, result_species$g_df)

hist(result_species$g_df$gamma)

plot_species <- topic_membership_res_species$plot + 
  stat_compare_means(
    comparisons = list(c("Abscess", "Plaque")),
    method = "wilcox.test",
    label.y = c(1), 
    vjust = .45,
    label = "p.signif")+ 
    ylim(0, 1.1)

plot_species <- plot_species + scale_x_discrete(labels = c("Abscess" = "A", "Plaque" = "P")) + 
  theme (
    axis.text=element_text(size=13, color="black"),
  )
plot_species

ggsave("output/Topic_Model_Topic_Membership_by_Type_Species.png", plot_species, width = 7.4, height = 2.5, dpi=600)
```

## Fig 7E: Plot beta
```{r, fig.height=3, fig.width=25}

### Plot beta
#Plot beta, or the numbers that are assigned to each word in a topic. If the beta score is higher, the word matters more in that topic.
my_colors <- c("1" = "grey", "2" = "#3185FC", "3" = "#FF495C", "4"= "#3185FC", "5" = "grey", "6" = "grey")


result_species$b_df$topic <- factor(result_species$b_df$topic, levels = c("2", "4", "3", "1", "5", "6"))

#my_colors <- c("1" = "#3185FC", "2" = "#3185FC", "3" = "#FF495C", "4" ="#FF495C" , "5" = "grey")
p <- plot_beta(result_species, 10, result_species$b_df,  fill_colors = my_colors)
p 

ggsave("output/Topic_Model_Beta_Barplots_Species.png", p, width = 25, height = 3, dpi=600)
```

## Fig 7B: Heatmap of gamma scores 
```{r, fig.height=2.3, fig.width=9}
heatmap_gamma_species <- heatmap_gamma(result_species, "Type", result_species$g_df)

png("output/Topic_Model_Gamma_Heatmap_Species.png",
    width = 9, height = 2.3, units = "in", res = 600)

ComplexHeatmap::draw(heatmap_gamma_species)  # <- explicitly use this one

dev.off()
```


## Fig 7D: UMAP of gamma scores
```{r fig.height=6, fig.width=6}
UMAP_res_sp <- plot_gamma_umap(result_species, "Type", colors_all, result_species$g_df)
UMAP_res_sp$umap

ggsave("output/Topic_Model_Gamma_UMAP_Species.png", UMAP_res_sp$umap, width = 5, height = 4, dpi=600)
```

```{r, fig.width=3, fig.height=3}
UMAP_res_sp$sil_plot

ggsave("output/Topic_Model_Gamma_UMAP_Silhouette_Species.png", UMAP_res_sp$sil_plot, width = 3, height = 3, dpi=600)
```



## Fig S6, S7: Relab heatmaps by topic
```{r, fig.height=2, fig.width=7}
topic1_res_species <- heatmap_by_topic(result_species$b_df, ps_fs_species_csv, n_top = 10, topic = 1)

png("output/Topic_Model_RelAb_Heatmap_Topic1_Species.png",
    width = 6.5, height = 2.5, units = "in", res = 600)

ComplexHeatmap::draw(topic1_res_species$heatmap)

dev.off()

topic1_res_species$heatmap
```

```{r, fig.height=2, fig.width=7.5}
topic2_res_species <- heatmap_by_topic(result_species$b_df, ps_fs_species_csv, n_top = 12, topic = 2)

png("output/Topic_Model_RelAb_Heatmap_Topic2_Species.png",
    width = 6.5, height = 2.5, units = "in", res = 600)

ComplexHeatmap::draw(topic2_res_species$heatmap)  # <- explicitly use this one

dev.off()

topic2_res_species$heatmap

```

```{r, fig.height=2.5, fig.width=6.5}
topic3_res_species <- heatmap_by_topic(result_species$b_df, ps_fs_species_csv, n_top = 12, topic = 3)

png("output/Topic_Model_RelAb_Heatmap_Topic3_Species.png",
    width = 6.5, height = 2.5, units = "in", res = 600)

ComplexHeatmap::draw(topic3_res_species$heatmap)  # <- explicitly use this one

dev.off()

topic3_res_species$heatmap
```

```{r, fig.height=2, fig.width=7}
topic4_res_species <- heatmap_by_topic(result_species$b_df, ps_fs_species_csv, n_top = 10, topic = 4)

png("output/Topic_Model_RelAb_Heatmap_Topic4_Species.png",
    width = 6.5, height = 2.5, units = "in", res = 600)

ComplexHeatmap::draw(topic4_res_species$heatmap)  # <- explicitly use this one

dev.off()

topic4_res_species$heatmap

```

```{r, fig.height=2, fig.width=7}
topic5_res_species <- heatmap_by_topic(result_species$b_df, ps_fs_species_csv, n_top = 10, topic = 5)


png("output/Topic_Model_RelAb_Heatmap_Topic5_Species.png",
    width = 6.5, height = 2.5, units = "in", res = 600)

ComplexHeatmap::draw(topic5_res_species$heatmap)

dev.off()

topic5_res_species$heatmap
```

```{r, fig.height=2, fig.width=7}
topic6_res_species <- heatmap_by_topic(result_species$b_df, ps_fs_species_csv, n_top = 10, topic = 6)


png("output/Topic_Model_RelAb_Heatmap_Topic6_Species.png",
    width = 6.5, height = 2.5, units = "in", res = 600)

ComplexHeatmap::draw(topic6_res_species$heatmap)

dev.off()

topic6_res_species$heatmap
```


# Session Info
```{r}
sessionInfo()
```

